### 2023-11-14 draft_systoles.tex
regulator*cent_index <= log^2 p * (p-1)

\documentclass[a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage[english]{babel}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage{hyperref} % clickable links in the document
\usepackage{enumitem} % (a), (b), etc labels

\usepackage{xcolor} % support of \red text
\newcommand{\red}{\textcolor{red}} % support of \red text

\newtheorem{Thm}{Theorem}[section]
\newtheorem{Lem}[Thm]{Lemma}
\newtheorem{Def}[Thm]{Definition}
\newtheorem{Cor}[Thm]{Corollary}
\newtheorem{Prop}[Thm]{Proposition}
\newtheorem{Rem}[Thm]{Remark}
\newtheorem{Ex}[Thm]{Example}
\newtheorem{Cla}[Thm]{Claim}
\newtheorem{Exer}[Thm]{Exercise}
\newtheorem{Prob}[Thm]{Problem}

\newcommand{\embeds}{\hookrightarrow}
\newcommand{\projects}{\twoheadrightarrow}
\newcommand{\eqdef}{\stackrel{\mathrm{def}}{=}}

\DeclareMathOperator{\assign}{\coloneqq}        % Assingment operator
\newcommand{\R}{\mathbb{R}}        % Real
\newcommand{\Q}{\mathbb{Q}}        % rationals
\newcommand{\C}{\mathbb{C}}        % Complex
\renewcommand{\P}{\mathbb{P}}        % Projective
\newcommand{\A}{\mathbb{A}}        % Affine
\renewcommand{\O}{\mathcal{O}}        % Structure sheaf
\newcommand{\Id}{\mathbf{Id}}        % Identity
\newcommand{\SL}{\mathbf{SL}_3(\mathbb{Z})}        % SL_3(Z)
\newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
\newcommand{\SLp}{\mathbf{SL}_3(p)}        % SL_3(p)
% \newcommand{\SLp}{\Gamma_p}        % a congruence subgroup

\DeclareMathOperator{\Spec}{Spec}        % Spec
\DeclareMathOperator{\Proj}{Proj}        % Proj
\DeclareMathOperator{\Reg}{Reg}        % Regulator
\DeclareMathOperator{\disc}{disc}        % discriminant

\begin{document}
\title{A sample computation}	
\author{Altan Erdnigor}
\maketitle

\section{Notation}
\begin{itemize}
\item $p$ a prime number.
\item $\SL$ the special linear group over $\mathbb{Z}$.
\item $\SLp$ the $p$th congruence subgroup of $\SL$.
\item $Z_G(x)$ the centralizer of $x \in G$.
\item 
\begin{equation}
\label{matrix_small}
A = 
\begin{pmatrix}
0 & 0 & 1 \\
1 & 0 & p+1 \\
0 & 1 & p
\end{pmatrix}
\in \SL
\end{equation}
\item 
\begin{equation}
\label{matrix_big}
\tilde A = \Id + p A =
\begin{pmatrix}
1 & 0 & p \\
p & 1 & p + p^2 \\
0 & p & 1 + p^2
\end{pmatrix}
\in \SLp
\end{equation}

\item If $C$ is a matrix, $\chi_C(\lambda) \assign \det(\lambda \Id - C)$ is the characteristics polynomial.
\item $ f(t) \assign \chi_A(t) = t^3 - p t^2 - (p + 1) t - 1$.

\end{itemize}

\section{Intro}
In this note we find an infinite series of primes  
\begin{equation}
	\label{the_primes_sequence}
59, 101, 167, 173, 211, 223, 271, 307, 317, 347, 449, 463, 593, 599, 607, 691, 719, 809, 821, 829, 853, \ldots
\end{equation}
with the following properties:
\begin{enumerate}
\item
The regularor of $\tilde A$
grows as $\approx \ln^2(p)$.

That is, $\Reg ( \Q(\alpha)/\Q ) \approx \ln^2(p) $ for $\alpha$ a root of $\chi_{\tilde A}$.

\item
The index of the centralizers
\[
	[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)] < 3p
\] 
is bounded by $3p$.
\end{enumerate}
Combining them both one gets something about systoles which is of interest of geometric group theory?
\section{Regulator}
We refer to Keith Conrad's write-up on Dirichlet's unit theorem and regulators \cite{conraddirichlet} for the definitions.
The current proof mimics the proof of Theorem 5.12 of Conrad.

Notice that 
\begin{multline}
\chi_{\tilde A}(\lambda) 
= \det(\lambda \Id - \tilde A) \\
= \det(\lambda \Id - (\Id + p A))
= \det((\lambda - 1) \Id - p A)\\
= p^3 \det(\frac{\lambda - 1}p \Id - A)
= p^3 \chi_A(\frac{\lambda - 1}p)
.\end{multline}
Hence adding the root of $\chi_{A}$ or $\chi_{\tilde A}$ result in the same field; therefore we reduce to showing that 
$\Reg ( \Q(\alpha)/\Q ) \approx \ln^2(p) $ for $\alpha$ a root of $\chi_{A}(t) = f(t) = t^3 - (p t^2 + (p + 1) t + 1) $.

\begin{Lem}
$\Q(\alpha)/\Q$ is totally real of degree $3$ for primes $p \ne 2$.
\end{Lem}
\begin{proof}
$f(t)$ is irreducible over $\Q$; indeed, by the real roots theorem it's sufficient to check $\pm 1$:
$$f(1) = - 2 p - 1, f(-1) = - 1 .$$

A simple computation shows that the discriminant of $f(t)$ equals 
\[
	\disc_f(p) = p^4 + 2 p^3 - 5 p^2 - 6 p - 23 
.\]
If $p \ge 3$ the discriminant  $\disc_f(p) > 0$ is positive.
Therefore the cubic extension is totally real.
\end{proof}

\begin{Prop}
	\label{prop_multiplicative_group_structure}
$\mathbb{Z}[\alpha]^{*} = \{ \pm \alpha^a (\alpha + 1)^b \mid a, b \in \mathbb{Z} \} $.
\end{Prop}
Note that $\alpha, \alpha + 1$ are not necessarily fundamental units in $\Q(\alpha)/\Q$ as we don't claim that the ring of integers of $\Q(\alpha) / \Q$ coincides with $\mathbb{Z}[\alpha]$.
\begin{proof}
Note that $f(\alpha) = 0$ implies 
\begin{equation}
\label{equation_unit}
\alpha (\alpha + 1) (\alpha - p) = 1
\end{equation}
It shows that $\alpha, \alpha + 1$ are indeed units.

Let $\alpha_1 > \alpha_2 > \alpha_3 \in \R$ be the three different roots of $f$.
We shall compute them approximately using \eqref{equation_unit}.
\begin{gather*}
\alpha_1 = p + 1 + O\left(\frac{1}{p^3}\right), \\
\alpha_2 = - \frac{1}{ p } + O\left(\frac{1}{p^4}\right), \\
\alpha_3 = - 1 + \frac{1}{ p } + O\left(\frac{1}{p^3}\right)
.\end{gather*}
\begin{Rem}
A computation shows that for $p = 10000$
we have
 \[
\alpha_1 = 10001.000000009999 ,
\alpha_2 = -0.00010000000000010001 ,
\alpha_3 = -0.9999000099970006 
.\] 
It is not important that $p$ is not a prime in this case as the estimate works for any sufficiently large $p$.
\end{Rem}

By the definition of the regulator we have
\begin{multline}
\Reg(\alpha, \alpha + 1) = 
\begin{vmatrix}
\ln |\alpha_1| 		& \ln |\alpha_2| \\
\ln |\alpha_1 + 1| 	& \ln |\alpha_2 + 1|
\end{vmatrix} 
\\
\approx
\begin{vmatrix}
\ln |p + 1| 	& \ln |\frac{-1}{p}| \\
\ln |p + 2| 	& \ln |\frac{-1}{p} + 1|
\end{vmatrix}
=
\ln(p + 1)
(\ln(p - 1) - \ln(p))
+
\ln(p + 2) \ln(p)
\\
=
\ln(p + 1) \ln(p - 1) - \ln(p + 1) \ln(p) + \ln(p + 2) \ln(p)
.\end{multline}
Therefore $ \Reg(\alpha, \alpha + 1) > 0 $ for all prime $p$.

Hence $\alpha, \alpha + 1$ are independent units.

By Corollary 5.9 from Conrad it is sufficient to check
\[
\frac{16 \Reg(\alpha, \alpha + 1) }
{(\ln(\disc_f/4))^2} 
< 2
.\] 
Substituting, we obtain
\[
\frac{16 \Reg(\alpha, \alpha + 1) }
{(\ln(\disc_f/4))^2} 
\approx
\frac{16 
\left(\ln(p + 1) \ln(p - 1) - \ln(p + 1) \ln(p) + \ln(p + 2) \ln(p)\right)
}
{(\ln(
(p^4 + 2 p^3 - 5 p^2 - 6 p - 23)
/4))^2} 
.\] 
Asymptotically, the latter equals
\[
\stackrel{p \to \infty}{\longrightarrow}
\frac{16 
\ln(p)^2
}
{(\ln(
p^4 ))^2} = 1
.\] 
Therefore it is $< 2$ for big enough $p$, QED.
\end{proof}
\begin{Rem}
One can do the estimate more carefully, but for now postpone it. 
Note that the sequence \ref{the_primes_sequence} begins from $p = 59$, thus we can restrict onto $p \ge 59$.
\end{Rem}
\begin{Rem}
We just proved that the regulator is approximately 
\[
\ln(p + 1) \ln(p - 1) - \ln(p + 1) \ln(p) + \ln(p + 2) \ln(p)
,\] 
which is close to $\ln^2 p$ we wanted from the beginning.
\end{Rem}

\section{Centralizers}
\begin{Prop}
	\label{proposition_centralizer_in_sl}
The centralizer of $\tilde A$ in $\SL$
is generated by $A, A+\Id$.
$$Z_{\SL}(\tilde A) = \{ \pm A^a (A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
\end{Prop}
\begin{proof}
Since $\tilde A$ is regular, its centralizer in $ \Mat $ is $\C \left< \Id, A, A^2 \right> $.
Now, 
\[
\C \left< \Id, A, A^2 \right> \cap \SL 
\subset \mathbb{Z} 
\left< \Id, A, A^2 \right>
.\] 
Indeed,
\begin{equation}
\Id = 
\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{pmatrix}, 
A = 
\begin{pmatrix}
0 & 0 & 1 \\
1 & 0 & p+1 \\
0 & 1 & p
\end{pmatrix}, 
A^2 = 
\begin{pmatrix}
0 & 1 & p \\
0 & p+1 & p^2+p+1 \\
1 & p & p^2+p+1
\end{pmatrix}, 
\end{equation}
Considering the first matrix column we see that if a complex combination has integer coefficients, it is in fact integer combination.

Moreover, 
the centralizer of $\tilde A$ is a group, therefore it lies inside the multiplicative group of $\mathbb{Z}[A]$
\[
Z_{\SL}(\tilde A)  \subset 
\C \left< \Id, A, A^2 \right> \cap \SL 
\subset \mathbb{Z}[A]^*
.\] 
There is an isomorphism of $\mathbb{Z}$-algebras $\mathbb{Z}[A] \simeq \mathbb{Z}[x]/(f(x)) = \mathbb{Z}[\alpha]$.
Applying Proposition \ref{prop_multiplicative_group_structure} end the proof
\[
Z_{\SL}(\tilde A) 
\subset \mathbb{Z}[A]^* =\{ \pm A^a (A + \Id)^b \mid a, b \in \mathbb{Z} \}.\] 
\end{proof}

We are to study the centralizer of $\tilde A$ in $\SLp$.
\[
Z_{\SLp}(\tilde A) \subset 
Z_{\SL}(\tilde A) \cong \mathbb{Z}^2
.\] 
In general this subgroup can be difficult to describe; 
that leads to the following
\begin{Def}
	A prime number $p$ has property $Ch$ if $t^3 - t - 1 \in \mathbb F_p[t]$ splits completely, that is  
\[
t^3 - t - 1 = (t - \beta_1)(t - \beta_2)(t - \beta_3)
,\] 
for some $\beta_1, \beta_2, \beta_3 \in \mathbb F_p$.
\end{Def}
The sequence \eqref{the_primes_sequence} is exactly the sequence of primes with the property $Ch$.
Why is it infinite?
\begin{Prop}
The density of primes with the property $Ch$ is $\frac{1}{6}$; in particular, there are infinite number of such primes.
\end{Prop}
\begin{proof}
Let $E/ \Q$ be the splitting field of $g(t) = t^3 - t - 1$.
The discriminant of $g$ equals $-23$, hence $E / \Q$ is of degree $6$.

Recall that a prime $p$ in $\Q$ is said to \emph{split completely} iff 
\[
p \mathbb{Z} = \gamma_1 \gamma_2 \gamma_3 \gamma_4 \gamma_5 \gamma_6
,\] 
for $\gamma_i$ primes in $\O_E$.
\red{$p$ splits completely iff it satisfies $Ch$}.
\red{feels close to Cox \cite{cox2022primes} Proposition 5.11}.

By Chebotarev's density theorem the density of primes splitting completely in $E$ is one over the degree of the extension.
\end{proof}

\begin{Prop}
Let prime $p$ satisfy $Ch$.

The centralizer $Z_{\SLp}(\tilde A)$ contains $\tilde A$ and $(A + \Id)^{p - 1}$.
\end{Prop}
\begin{proof}
The only thing to prove is that 
$(A + \Id)^{p - 1} \in \SLp$.
It suffices to show 
\[
	%(\alpha + 1)^{p-1} - 1 \in p \mathbb{Z}[\alpha] ~ ~  (or \in p \O_{\Q(\alpha)}  ? )
	(x + 1)^{p - 1} - 1 \in (p, x^3 - x - 1)
.\] 
The polynomial vanishes at all but one residues from $\mathbb F_p$, namely at $\mathbb F_p \setminus \{ -1 \} $.

By the $Ch$ property $x^3 - x - 1$ has three distinct roots in $\mathbb F_p$; clearly $-1$ is not one of them.

Therefore the polynomial $ x^3 - x - 1 $ divides $ (x + 1)^{p - 1} - 1 $ in $\mathbb F_p[x]$, and we are done.
\end{proof}

\begin{Thm}
	The index of the centralizers divides $3 (p - 1) $. 
\[
	[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)] | 3 (p-1)
.\] 
In particular, it is smaller than $3 p$.
\end{Thm}
\begin{proof}
Identify $Z_{\SL}(\tilde A) \cong \mathbb{Z}/2\mathbb{Z} \oplus \mathbb{Z}^2$ as in Proposition \ref{proposition_centralizer_in_sl}.

Observe that 
\[
1 + p \alpha = \alpha^3 (\alpha + 1)^{-1} 
.\] 
Indeed, multiplying both sides by $(\alpha + 1)$ we obtain
\[
1 + (p + 1) \alpha + \alpha^2 = \alpha^3 
,\] 
which holds true by the definition of $\alpha$.

By the previous Proposition 
$Z_{\SLp}(\tilde A)$ contains $\binom{3}{-1}, \binom{0}{p-1}$;
Clearly, the index
\[
\mathbb{Z} \left< \binom{3}{-1}, \binom{0}{p-1}\right> \subset \mathbb{Z}^2
,\] 
equals $3 (p-1)$ and the index 
$ [Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$
has to divide it.
\end{proof}

\section{Fantasies}
\subsection{An exercise in generating functions}

Let 
$$
f(x) = x^3 - a x^2 + b x - c
.$$ 
be a qubic polynomial.
Let $\alpha_1, \alpha_2, \alpha_3$ to be its roots.

We have
$$
t^3 f(t^{-1}) = 1 - a t + b t^2 - c t^3 = (1 - t \alpha_1)(1 - t \alpha_2)(1 - t \alpha_3)
.$$

The goal is to prove
\begin{Prop}
\label{prop_generating_function_roots_powers}
Define a generating series
\[
F(t, q) \assign \sum_{i = 0}^\infty 
(1 - t \alpha_1^i)(1 - t \alpha_2^i)(1 - t \alpha_3^i) q^i
.\] 
Then we have 
\[
F(t, q) =
\frac{1}{1 - q}
- t \frac{ 3 - 2 a q + b q^2 }{1 - a q + b q^2 - c q^3}
+ t^2 \frac{ 3 - 2 b q + a c q^2 }{ 1 - b q + a c q^2 - c^2 q^3 }
- t^3 \frac{1}{ 1 - c q}
.\] 
\end{Prop}
\begin{proof}
A direct computation.
\begin{multline}
F(t, q) 
= \sum_{i = 0}^\infty 
(1 - t \alpha_1^i)(1 - t \alpha_2^i)(1 - t \alpha_3^i) q^i
\\ = 
\sum_{i = 0}^\infty  q^i
- t \sum_{i = 0}^\infty (\alpha_1^i + \alpha_2^i +  \alpha_3^i) q^i
+ t^2 \sum_{i = 0}^\infty (\alpha_2^i\alpha_3^i + \alpha_1^i\alpha_3^i + \alpha_1^i\alpha_2^i) q^i
- t^3 \sum_{i = 0}^\infty (\alpha_1^i \alpha_2^i \alpha_3^i) q^i
\\ = 
\frac{1}{1 - q}
- t \left( \frac{1}{1 - \alpha_1 q} + \frac{1}{1 - \alpha_2 q}  + \frac{1}{1 - \alpha_3 q}\right) 
\\ + t^2  \left( \frac{1}{1 - \alpha_2 \alpha_3 q} + \frac{1}{1 - \alpha_1 \alpha_3 q}  + \frac{1}{1 - \alpha_1 \alpha_2 q}\right) 
- t^3 \frac{1}{1 - \alpha_1 \alpha_2 \alpha_3 q}
\\ = 
\frac{1}{1 - q}
- t \frac{{(1 - \alpha_2 q)( 1 - \alpha_3 q) + (1 - \alpha_1 q)( 1 - \alpha_3 q) + (1 - \alpha_1 q)( 1 - \alpha_2 q) }
}{(1 - \alpha_1 q)( 1 - \alpha_2 q)(1 - \alpha_3 q)}
\\ + t^2 \frac{{(1 - \alpha_1\alpha_3 q)( 1 - \alpha_1\alpha_2 q) + (1 - \alpha_2\alpha_3 q)( 1 - \alpha_1\alpha_2 q) + (1 - \alpha_2\alpha_3 q)( 1 - \alpha_1\alpha_3 q) }
}{(1 - \alpha_2\alpha_3 q)( 1 - \alpha_1\alpha_3 q)(1 - \alpha_1\alpha_2 q)}
\\ - t^3 \frac{1}{1 - \alpha_1 \alpha_2 \alpha_3 q}
\\ = 
\frac{1}{1 - q}
- t 
\frac{
	3 - 2 (\alpha_1 + \alpha_2 + \alpha_3) q + (\alpha_2\alpha_3 + \alpha_1\alpha_3 + \alpha_1\alpha_2) q^2
}{q^3 f(q^{-1} )}
\\ + t^2 \frac{
	3 - 2 (\alpha_2 \alpha_3 + \alpha_1 \alpha_3 + \alpha_1 \alpha_2 ) q + (\alpha_1 + \alpha_2 + \alpha_3) \alpha_1 \alpha_2 \alpha_3  q^2
}{
	1 - (\alpha_2 \alpha_3 + \alpha_1 \alpha_3 + \alpha_1 \alpha_2 ) q + (\alpha_1 + \alpha_2 + \alpha_3) \alpha_1 \alpha_2 \alpha_3 q^2 - \alpha_1^2 \alpha_2^2 \alpha_3^2 q^3
}
\\ - t^3 \frac{1}{1 - \alpha_1 \alpha_2 \alpha_3 q}
\\ = 
\frac{1}{1 - q}
- t \frac{ 3 - 2 a q + b q^2 }{1 - a q + b q^2 - c q^3}
+ t^2 \frac{ 3 - 2 b q + a c q^2 }{ 1 - b q + a c q^2 - c^2 q^3 }
- t^3 \frac{1}{1 - c q}
\end{multline}
\end{proof}

Apply Proposition  
\ref{prop_generating_function_roots_powers}
to 
$$g(t) = t^3 - t - 1.$$
Therefore
\[
G(t, q) = 
\frac{1}{1 - q}
- t \frac{ 3 - q^2 }{1 - q^2 - q^3}
+ t^2 \frac{ 3 + 2 q }{ 1 + q - q^3 }
- t^3 \frac{1}{1 - q}
.\]

\bibliography{mybib}
\bibliographystyle{plain}

\end{document}

### 2024-02-01 draft_systoles.tex
save November 14 version

@@
 0 & 0 & 1 \\
-1 & 0 & p+1 \\
-0 & 1 & p
+1 & 0 & p+2 \\
+0 & 1 & 2p
 \end{pmatrix}
@@
 1 & 0 & p \\
-p & 1 & p + p^2 \\
-0 & p & 1 + p^2
+p & 1 & 2 p + p^2 \\
+0 & p & 1 + 2 p^2
 \end{pmatrix}
@@
 \item If $C$ is a matrix, $\chi_C(\lambda) \assign \det(\lambda \Id - C)$ is the characteristics polynomial.
-\item $ f(t) \assign \chi_A(t) = t^3 - p t^2 - (p + 1) t - 1$.
+\item $ f(t) \assign \chi_A(t) = t^3 - 2p t^2 - (p + 2) t - 1$.
 
@@
 \section{Intro}
-In this note we find an infinite series of primes  
+Recall Lucas primes (see \url{https://en.wikipedia.org/wiki/Lucas_number#Lucas_primes}, \url{https://t5k.org/top20/page.php?id=48})
 \begin{equation}
 	\label{the_primes_sequence}
-59, 101, 167, 173, 211, 223, 271, 307, 317, 347, 449, 463, 593, 599, 607, 691, 719, 809, 821, 829, 853, \ldots
+2, 3, 7, 11, 29, 47, 199, 521, 2207, 3571, 9349, 3010349, 54018521, 370248451, 6643838879, \ldots
 \end{equation}
-with the following properties:
+In this notes we establish the following results:
 \begin{enumerate}
@@
 Hence adding the root of $\chi_{A}$ or $\chi_{\tilde A}$ result in the same field; therefore we reduce to showing that 
-$\Reg ( \Q(\alpha)/\Q ) \approx \ln^2(p) $ for $\alpha$ a root of $\chi_{A}(t) = f(t) = t^3 - (p t^2 + (p + 1) t + 1) $.
+$\Reg ( \Q(\alpha)/\Q ) \approx \ln^2(p) $ for $\alpha$ a root of $\chi_{A}(t) = f(t) = t^3 - (2p t^2 + (p + 2) t + 1) $.
 
@@
 $f(t)$ is irreducible over $\Q$; indeed, by the real roots theorem it's sufficient to check $\pm 1$:
-$$f(1) = - 2 p - 1, f(-1) = - 1 .$$
+$$f(1) = - 3 p - 2, f(-1) = - p .$$
 
@@
 \[
-	\disc_f(p) = p^4 + 2 p^3 - 5 p^2 - 6 p - 23 
+	\disc_f(p) = 4 p^4 - 12 p^3 + 4 p^2 - 24 p + 5
 .\]
-If $p \ge 3$ the discriminant  $\disc_f(p) > 0$ is positive.
+If $p > 4$ the discriminant  $\disc_f(p) > 0$ is positive.
 Therefore the cubic extension is totally real.
@@
 	\label{prop_multiplicative_group_structure}
-$\mathbb{Z}[\alpha]^{*} = \{ \pm \alpha^a (\alpha + 1)^b \mid a, b \in \mathbb{Z} \} $.
+$\mathbb{Z}[\alpha]^{*} = \{ \pm \alpha^a (2\alpha + 1)^b \mid a, b \in \mathbb{Z} \} $.
 \end{Prop}
@@
 Note that $f(\alpha) = 0$ implies 
-\begin{equation}
-\label{equation_unit}
-\alpha (\alpha + 1) (\alpha - p) = 1
-\end{equation}
-It shows that $\alpha, \alpha + 1$ are indeed units.
+\begin{gather}
+\alpha (\alpha^2 - 2p \alpha - (p + 2)) = 1
+\\
+\label{equation_units}
+(1 + 2 \alpha) (1 + p \alpha) = \alpha^3
+\end{gather}
+It shows that $\alpha, 1 + 2 \alpha$ are indeed units.
 
 Let $\alpha_1 > \alpha_2 > \alpha_3 \in \R$ be the three different roots of $f$.
-We shall compute them approximately using \eqref{equation_unit}.
+We shall compute them approximately.
 \begin{gather*}
-\alpha_1 = p + 1 + O\left(\frac{1}{p^3}\right), \\
+\alpha_1 = 2 p + \frac{1}{2} + O\left(\frac{1}{p}\right), \\
 \alpha_2 = - \frac{1}{ p } + O\left(\frac{1}{p^4}\right), \\
-\alpha_3 = - 1 + \frac{1}{ p } + O\left(\frac{1}{p^3}\right)
+\alpha_3 = - \frac{1}{2} + O\left(\frac{1}{p}\right)
 .\end{gather*}
+% from sympy import *
+% [ins] In [2]: x = symbols('x')
+% [ins] In [3]: p = 10000
+% [ins] In [12]: f = x**3 - 2 * p * x**2 - (p + 2) * x - 1
+% [ins] In [13]: f = Poly(f)
+% [ins] In [16]: nroots(f)
+% Out[16]: [-0.499987498124648, -0.000100000000000100, 20000.5000874981]
 \begin{Rem}
+
 A computation shows that for $p = 10000$
@@
  \[
-\alpha_1 = 10001.000000009999 ,
-\alpha_2 = -0.00010000000000010001 ,
-\alpha_3 = -0.9999000099970006 
+\alpha_1 = 20000.5000874981 ,
+\alpha_2 = -0.000100000000000100 ,
+\alpha_3 = -0.499987498124648 
 .\] 
@@
 \ln |\alpha_1| 		& \ln |\alpha_2| \\
-\ln |\alpha_1 + 1| 	& \ln |\alpha_2 + 1|
+\ln |2 \alpha_1 + 1| 	& \ln |2 \alpha_2 + 1|
 \end{vmatrix} 
@@
 \begin{vmatrix}
-\ln |p + 1| 	& \ln |\frac{-1}{p}| \\
-\ln |p + 2| 	& \ln |\frac{-1}{p} + 1|
+\ln |2p + \frac{1}{2}| 	& \ln |\frac{-1}{p}| \\
+\ln |4 p + 2| 	& \ln |\frac{-2}{p} + 1|
 \end{vmatrix}
 =
-\ln(p + 1)
-(\ln(p - 1) - \ln(p))
+\ln(2p + \frac{1}{2})
+(\ln(p - 2) - \ln(p))
 +
-\ln(p + 2) \ln(p)
+\ln(4p + 2) \ln(p)
 \\
 =
-\ln(p + 1) \ln(p - 1) - \ln(p + 1) \ln(p) + \ln(p + 2) \ln(p)
+\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
 .\end{multline}
@@
 \approx
-\frac{16 
-\left(\ln(p + 1) \ln(p - 1) - \ln(p + 1) \ln(p) + \ln(p + 2) \ln(p)\right)
-}
+\frac{16 (
+\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
+) }
 {(\ln(
@@
 One can do the estimate more carefully, but for now postpone it. 
-Note that the sequence \ref{the_primes_sequence} begins from $p = 59$, thus we can restrict onto $p \ge 59$.
 \end{Rem}
@@
 \[
-\ln(p + 1) \ln(p - 1) - \ln(p + 1) \ln(p) + \ln(p + 2) \ln(p)
+\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
 ,\] 
@@
 is generated by $A, A+\Id$.
-$$Z_{\SL}(\tilde A) = \{ \pm A^a (A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
+$$Z_{\SL}(\tilde A) = \{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
 \end{Prop}
@@
 0 & 0 & 1 \\
-1 & 0 & p+1 \\
-0 & 1 & p
+1 & 0 & p+2 \\
+0 & 1 & 2p
 \end{pmatrix}, 
@@
 \begin{pmatrix}
-0 & 1 & p \\
-0 & p+1 & p^2+p+1 \\
-1 & p & p^2+p+1
+0 & 1 & 2p \\
+0 & p+1 & 2p^2+4p+1 \\
+1 & p & 4p^2+p+2
 \end{pmatrix}, 
@@
 Z_{\SL}(\tilde A) 
-\subset \mathbb{Z}[A]^* =\{ \pm A^a (A + \Id)^b \mid a, b \in \mathbb{Z} \}.\] 
+\subset \mathbb{Z}[A]^* =\{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \}.\] 
 \end{proof}
@@
 In general this subgroup can be difficult to describe; 
-that leads to the following
-\begin{Def}
-	A prime number $p$ has property $Ch$ if $t^3 - t - 1 \in \mathbb F_p[t]$ splits completely, that is  
-\[
-t^3 - t - 1 = (t - \beta_1)(t - \beta_2)(t - \beta_3)
-,\] 
-for some $\beta_1, \beta_2, \beta_3 \in \mathbb F_p$.
-\end{Def}
-The sequence \eqref{the_primes_sequence} is exactly the sequence of primes with the property $Ch$.
-Why is it infinite?
-\begin{Prop}
-The density of primes with the property $Ch$ is $\frac{1}{6}$; in particular, there are infinite number of such primes.
-\end{Prop}
-\begin{proof}
-Let $E/ \Q$ be the splitting field of $g(t) = t^3 - t - 1$.
-The discriminant of $g$ equals $-23$, hence $E / \Q$ is of degree $6$.
-
-Recall that a prime $p$ in $\Q$ is said to \emph{split completely} iff 
-\[
-p \mathbb{Z} = \gamma_1 \gamma_2 \gamma_3 \gamma_4 \gamma_5 \gamma_6
-,\] 
-for $\gamma_i$ primes in $\O_E$.
-\red{$p$ splits completely iff it satisfies $Ch$}.
-\red{feels close to Cox \cite{cox2022primes} Proposition 5.11}.
-
-By Chebotarev's density theorem the density of primes splitting completely in $E$ is one over the degree of the extension.
-\end{proof}
+that leads to cosidering \emph{Lucas primes}.
 
 \begin{Prop}
-Let prime $p$ satisfy $Ch$.
+Let $p$ be a Lucas prime.
+Let $k$ be the integer part of $\log_{\phi} p$ where $\phi$ is the golden ratio.
 
-The centralizer $Z_{\SLp}(\tilde A)$ contains $\tilde A$ and $(A + \Id)^{p - 1}$.
+The centralizer $Z_{\SLp}(\tilde A)$ contains $\tilde A$ and $A^{4k}$.
 \end{Prop}
@@
 The only thing to prove is that 
-$(A + \Id)^{p - 1} \in \SLp$.
-It suffices to show 
+$A^k \in \SLp$.
+
+It suffices to show that the eigenvalues of $A \pmod p$ are $4k$-th roots of unity.
+Computing
 \[
-	%(\alpha + 1)^{p-1} - 1 \in p \mathbb{Z}[\alpha] ~ ~  (or \in p \O_{\Q(\alpha)}  ? )
-	(x + 1)^{p - 1} - 1 \in (p, x^3 - x - 1)
+\chi_{A}(t) = f(t) \equiv t^3 - 2 t - 1 = (t + 1) (t^2 - t - 1) \pmod p
+,\] 
+shows that it is left to work with the golden ratio in $\mathbb F_p$.
+
+By the definition of a Lucas number we have 
+\[
+p = \phi^k + \phi^{-k} 
+,\] 
+where $\phi$ is a root of $x^2 - x - 1$.
+Taking it modulo $p$ we see that $\phi^k \pmod p$ satisfies 
+\[
+0 = y + y^{-1} 
+,\] 
+which implies 
+\[
+1 = y^4 \equiv \phi^{4k} \pmod p
 .\] 
-The polynomial vanishes at all but one residues from $\mathbb F_p$, namely at $\mathbb F_p \setminus \{ -1 \} $.
-
-By the $Ch$ property $x^3 - x - 1$ has three distinct roots in $\mathbb F_p$; clearly $-1$ is not one of them.
-
-Therefore the polynomial $ x^3 - x - 1 $ divides $ (x + 1)^{p - 1} - 1 $ in $\mathbb F_p[x]$, and we are done.
 \end{proof}
@@
 \begin{Thm}
-	The index of the centralizers divides $3 (p - 1) $. 
+	The index of the centralizers is bounded by $4 \log_{\phi} p$
 \[
-	[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)] | 3 (p-1)
+	[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)] \le 4 \log_{\phi} p
 .\] 
-In particular, it is smaller than $3 p$.
+In particular, grows as $O(\ln p)$.
 \end{Thm}
@@
 
-Observe that 
+Observe that by \eqref{equation_units} we have
 \[
-1 + p \alpha = \alpha^3 (\alpha + 1)^{-1} 
+1 + p \alpha = \alpha^3 (2 \alpha + 1)^{-1} 
 .\]
-Indeed, multiplying both sides by $(\alpha + 1)$ we obtain
-\[
-1 + (p + 1) \alpha + \alpha^2 = \alpha^3 
-,\] 
-which holds true by the definition of $\alpha$.
 
@@
 \[
-\mathbb{Z} \left< \binom{3}{-1}, \binom{0}{p-1}\right> \subset \mathbb{Z}^2
+\mathbb{Z} \left< \binom{3}{-1}, \binom{4k}{0}\right> \subset \mathbb{Z}^2
 ,\] 
-equals $3 (p-1)$ and the index 
+equals $4k$ and the index 
 $ [Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$
@@
 \end{proof}
-
-\section{Fantasies}
-\subsection{An exercise in generating functions}
-
-Let 
-$$
-f(x) = x^3 - a x^2 + b x - c
-.$$ 
-be a qubic polynomial.
-Let $\alpha_1, \alpha_2, \alpha_3$ to be its roots.
-
-We have
-$$
-t^3 f(t^{-1}) = 1 - a t + b t^2 - c t^3 = (1 - t \alpha_1)(1 - t \alpha_2)(1 - t \alpha_3)
-.$$
-
-The goal is to prove
-\begin{Prop}
-\label{prop_generating_function_roots_powers}
-Define a generating series
-\[
-F(t, q) \assign \sum_{i = 0}^\infty 
-(1 - t \alpha_1^i)(1 - t \alpha_2^i)(1 - t \alpha_3^i) q^i
-.\] 
-Then we have 
-\[
-F(t, q) =
-\frac{1}{1 - q}
-- t \frac{ 3 - 2 a q + b q^2 }{1 - a q + b q^2 - c q^3}
-+ t^2 \frac{ 3 - 2 b q + a c q^2 }{ 1 - b q + a c q^2 - c^2 q^3 }
-- t^3 \frac{1}{ 1 - c q}
-.\] 
-\end{Prop}
-\begin{proof}
-A direct computation.
-\begin{multline}
-F(t, q) 
-= \sum_{i = 0}^\infty 
-(1 - t \alpha_1^i)(1 - t \alpha_2^i)(1 - t \alpha_3^i) q^i
-\\ = 
-\sum_{i = 0}^\infty  q^i
-- t \sum_{i = 0}^\infty (\alpha_1^i + \alpha_2^i +  \alpha_3^i) q^i
-+ t^2 \sum_{i = 0}^\infty (\alpha_2^i\alpha_3^i + \alpha_1^i\alpha_3^i + \alpha_1^i\alpha_2^i) q^i
-- t^3 \sum_{i = 0}^\infty (\alpha_1^i \alpha_2^i \alpha_3^i) q^i
-\\ = 
-\frac{1}{1 - q}
-- t \left( \frac{1}{1 - \alpha_1 q} + \frac{1}{1 - \alpha_2 q}  + \frac{1}{1 - \alpha_3 q}\right) 
-\\ + t^2  \left( \frac{1}{1 - \alpha_2 \alpha_3 q} + \frac{1}{1 - \alpha_1 \alpha_3 q}  + \frac{1}{1 - \alpha_1 \alpha_2 q}\right) 
-- t^3 \frac{1}{1 - \alpha_1 \alpha_2 \alpha_3 q}
-\\ = 
-\frac{1}{1 - q}
-- t \frac{{(1 - \alpha_2 q)( 1 - \alpha_3 q) + (1 - \alpha_1 q)( 1 - \alpha_3 q) + (1 - \alpha_1 q)( 1 - \alpha_2 q) }
-}{(1 - \alpha_1 q)( 1 - \alpha_2 q)(1 - \alpha_3 q)}
-\\ + t^2 \frac{{(1 - \alpha_1\alpha_3 q)( 1 - \alpha_1\alpha_2 q) + (1 - \alpha_2\alpha_3 q)( 1 - \alpha_1\alpha_2 q) + (1 - \alpha_2\alpha_3 q)( 1 - \alpha_1\alpha_3 q) }
-}{(1 - \alpha_2\alpha_3 q)( 1 - \alpha_1\alpha_3 q)(1 - \alpha_1\alpha_2 q)}
-\\ - t^3 \frac{1}{1 - \alpha_1 \alpha_2 \alpha_3 q}
-\\ = 
-\frac{1}{1 - q}
-- t 
-\frac{
-	3 - 2 (\alpha_1 + \alpha_2 + \alpha_3) q + (\alpha_2\alpha_3 + \alpha_1\alpha_3 + \alpha_1\alpha_2) q^2
-}{q^3 f(q^{-1} )}
-\\ + t^2 \frac{
-	3 - 2 (\alpha_2 \alpha_3 + \alpha_1 \alpha_3 + \alpha_1 \alpha_2 ) q + (\alpha_1 + \alpha_2 + \alpha_3) \alpha_1 \alpha_2 \alpha_3  q^2
-}{
-	1 - (\alpha_2 \alpha_3 + \alpha_1 \alpha_3 + \alpha_1 \alpha_2 ) q + (\alpha_1 + \alpha_2 + \alpha_3) \alpha_1 \alpha_2 \alpha_3 q^2 - \alpha_1^2 \alpha_2^2 \alpha_3^2 q^3
-}
-\\ - t^3 \frac{1}{1 - \alpha_1 \alpha_2 \alpha_3 q}
-\\ = 
-\frac{1}{1 - q}
-- t \frac{ 3 - 2 a q + b q^2 }{1 - a q + b q^2 - c q^3}
-+ t^2 \frac{ 3 - 2 b q + a c q^2 }{ 1 - b q + a c q^2 - c^2 q^3 }
-- t^3 \frac{1}{1 - c q}
-\end{multline}
-\end{proof}
-
-Apply Proposition  
-\ref{prop_generating_function_roots_powers}
-to 
-$$g(t) = t^3 - t - 1.$$
-Therefore
-\[
-G(t, q) = 
-\frac{1}{1 - q}
-- t \frac{ 3 - q^2 }{1 - q^2 - q^3}
-+ t^2 \frac{ 3 + 2 q }{ 1 + q - q^3 }
-- t^3 \frac{1}{1 - q}
-.\]
 

### 2024-03-21 draft_systoles.tex
Minor corrections

@@
 \usepackage{hyperref} % clickable links in the document
+\usepackage{url} % clickable links in the document
 \usepackage{enumitem} % (a), (b), etc labels
@@
 \newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
-\newcommand{\SLp}{\mathbf{SL}_3(p)}        % SL_3(p)
-% \newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
+\newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
+\newcommand{\phip}{\phi_p}        % the golden ratio mod(p)
 
@@
 The index of the centralizers
-\[
-	[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)] < 3p
-\] 
-is bounded by $3p$.
+$ [Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A)] $
+grows as $O(\ln p)$ for $p$ a Lucas prime.
 \end{enumerate}
-Combining them both one gets something about systoles which is of interest of geometric group theory?
+This might be interesting.
 \section{Regulator}
@@
 \begin{proof}
-$f(t)$ is irreducible over $\Q$; indeed, by the real roots theorem it's sufficient to check $\pm 1$:
+$f(t)$ is irreducible over $\Q$; indeed, by the rational roots theorem it's sufficient to check $\pm 1$:
 $$f(1) = - 3 p - 2, f(-1) = - p .$$
@@
 \end{Prop}
-Note that $\alpha, \alpha + 1$ are not necessarily fundamental units in $\Q(\alpha)/\Q$ as we don't claim that the ring of integers of $\Q(\alpha) / \Q$ coincides with $\mathbb{Z}[\alpha]$.
+Note that $\alpha, 2\alpha + 1$ are not necessarily fundamental units in $\Q(\alpha)/\Q$ as we don't claim that the ring of integers of $\Q(\alpha) / \Q$ coincides with $\mathbb{Z}[\alpha]$.
 \begin{proof}
@@
 \begin{multline}
-\Reg(\alpha, \alpha + 1) = 
+\Reg(\alpha, 2\alpha + 1) = 
 \begin{vmatrix}
@@
 .\end{multline}
-Therefore $ \Reg(\alpha, \alpha + 1) > 0 $ for all prime $p$.
+Therefore $ \Reg(\alpha, 2\alpha + 1) > 0 $ for all prime $p$.
 
-Hence $\alpha, \alpha + 1$ are independent units.
+Hence $\alpha, 2\alpha + 1$ are independent units.
 
+It is left to prove that they are fundamental units in $\mathbb{Z}[\alpha]$.
 By Corollary 5.9 from Conrad it is sufficient to check
 \[
-\frac{16 \Reg(\alpha, \alpha + 1) }
+\frac{16 \Reg(\alpha, 2\alpha + 1) }
 {(\ln(\disc_f/4))^2} 
@@
 \[
-\frac{16 \Reg(\alpha, \alpha + 1) }
+\frac{16 \Reg(\alpha, 2\alpha + 1) }
 {(\ln(\disc_f/4))^2} 
@@
 The only thing to prove is that 
-$A^k \in \SLp$.
+$A^{4k}  \in \SLp$.
 
@@
 ,\] 
-shows that it is left to work with the golden ratio in $\mathbb F_p$.
+shows that it is left to work with the golden ratio in $\mathbb F_p$ which we denote by $\phip$.
+That is, $\phip \in \mathbb F_{p^2}$ satisfies $\phip^2 - \phip - 1 = 0$.
 
 By the definition of a Lucas number we have 
+\begin{equation}
+	\label{eq_lucas_prime_definition}
+p = \phi^k + (-\phi)^{-k} 
+\end{equation}
+where $\phi$ is a root of $x^2 - x - 1$.
+
+The RHS of \eqref{eq_lucas_prime_definition} being invariant under the change $\phi \to (-\phi)^{-1}$ manifests it as a symmetric polynomial in the roots of $x^2 - x - 1$, thus having a presentation
 \[
-p = \phi^k + \phi^{-k} 
+\phi^k + (-\phi)^{-k} = P( \phi, (-\phi)^{-1})
 ,\] 
-where $\phi$ is a root of $x^2 - x - 1$.
-Taking it modulo $p$ we see that $\phi^k \pmod p$ satisfies 
+where $P$ is a \emph{universal} polynomial.
+This observation justifies that the Equation \eqref{eq_lucas_prime_definition} can be taken modulo $p$ to have the form
 \[
-0 = y + y^{-1} 
+0 = \phip^k + (-\phip)^{-k} 
 ,\] 
@@
 \[
-1 = y^4 \equiv \phi^{4k} \pmod p
+1 = \phip^{4k} 
 .\] 
@@
 \[
-	[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)] \le 4 \log_{\phi} p
+	[Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A)] \le 4 \log_{\phi} p
 .\] 
-In particular, grows as $O(\ln p)$.
+In particular, it grows as $O(\ln p)$.
 \end{Thm}
@@
 By the previous Proposition 
-$Z_{\SLp}(\tilde A)$ contains $\binom{3}{-1}, \binom{0}{p-1}$;
+$Z_{\SLp}(\tilde A)$ contains $\binom{3}{-1}, \binom{4k}{0}$;
 Clearly, the index
@@
 ,\] 
-equals $4k$ and the index 
+equals $4k \approx 4 \log_{\phi} p$ and the index 
 $ [Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$

### 2024-04-01 systoles-SL3.tex
Add files via upload
Changed title, added coathors, added comment about relation to Artin's primitive rooot conjecture.

--- 2024-03-21 draft_systoles.tex
@@
 \begin{document}
-\title{A sample computation}	
-\author{Altan Erdnigor}
+\title{On $2$-systoles of congruence quotients of $\mathrm{SL}_3$.}	
+\author{Mikhail Belolipetsky, Altan Erdnigor, Shmuel Weinberger}
 \maketitle
@@
 \section{Intro}
-Recall Lucas primes (see \url{https://en.wikipedia.org/wiki/Lucas_number#Lucas_primes}, \url{https://t5k.org/top20/page.php?id=48})
+Recall the Lucas primes (see \url{https://en.wikipedia.org/wiki/Lucas_number#Lucas_primes}, \url{https://t5k.org/top20/page.php?id=48})
 \begin{equation}
@@
 .\end{multline}
-Therefore $ \Reg(\alpha, 2\alpha + 1) > 0 $ for all prime $p$.
+Therefore, $\Reg(\alpha, 2\alpha + 1) > 0 $ for all prime $p$.
 
-Hence $\alpha, 2\alpha + 1$ are independent units.
+Hence $\alpha$ and $2\alpha + 1$ are independent units.
 
@@
 .\] 
-Therefore it is $< 2$ for big enough $p$, QED.
+Therefore it is $< 2$ for big enough $p$.
 \end{proof}
@@
 The centralizer of $\tilde A$ in $\SL$
-is generated by $A, A+\Id$.
+is generated by $A, 2A+\Id$.
 $$Z_{\SL}(\tilde A) = \{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
@@
 
-\begin{Prop}
+\begin{Prop}\label{prop:lucas}
 Let $p$ be a Lucas prime.
@@
 where $P$ is a \emph{universal} polynomial.
-This observation justifies that the Equation \eqref{eq_lucas_prime_definition} can be taken modulo $p$ to have the form
+This observation justifies that equation \eqref{eq_lucas_prime_definition} can be taken modulo $p$ to have the form
 \[
@@
 \begin{Thm}
-	The index of the centralizers is bounded by $4 \log_{\phi} p$
+	If $p$ is a Lucuas prime, then the index of the centralizer is bounded by $4 \log_{\phi} p$:
 \[
@@
 
-By the previous Proposition 
+By the previous proposition 
 $Z_{\SLp}(\tilde A)$ contains $\binom{3}{-1}, \binom{4k}{0}$;
@@
 
+The observation that comes next is that the sequence of Lucas primes is very sparse.
+It is conjectured that there are infinitely many of them but even this is not known.
+In any respect, it is not hard to see that the number of Lucas primes less or equal than $N$ grows at most as fast as $O(\frac{1}{\log\log N})$.
+From a more general point of view we are interested in the order of the generators of $Z_{\SL}(\tilde A)$ mod $p$.
+One of the generators can be chosen as $\tilde{A}$ and is equal one mod $p$.
+We are left with the other generator.
+This can sometimes have very small order compared to $p$, like in the case with the Lucas primes considered in Proposition~\ref{prop:lucas}.
+However, in most of the cases the order will be $~p$ which implies that the index of the centralizer will be also of order $p$.
+
 \bibliography{mybib}

### 2024-04-01 systoles-SL3.tex
small changes

@@
 It is conjectured that there are infinitely many of them but even this is not known.
-In any respect, it is not hard to see that the number of Lucas primes less or equal than $N$ grows at most as fast as $O(\frac{1}{\log\log N})$.
+In any respect, it is not hard to see that the number of Lucas primes less or equal than $N$ grows at most as fast as $O(\frac{\log\log N}{N})$.
 From a more general point of view we are interested in the order of the generators of $Z_{\SL}(\tilde A)$ mod $p$.
-One of the generators can be chosen as $\tilde{A}$ and is equal one mod $p$.
+One of the generators can be chosen as $\tilde{A}$ and is equal to one mod $p$.
 We are left with the other generator.
 This can sometimes have very small order compared to $p$, like in the case with the Lucas primes considered in Proposition~\ref{prop:lucas}.
-However, in most of the cases the order will be $~p$ which implies that the index of the centralizer will be also of order $p$.
+However, in most of the cases the order will be $\sim p$ which implies that the index of the centralizer will be also of order $p$.
 

### 2024-05-26 draft_systoles.tex
test update draft_systoles.tex
add bounds, family, the unknown

@@
 \newcommand{\SL}{\mathbf{SL}_3(\mathbb{Z})}        % SL_3(Z)
+\newcommand{\GL}{\mathbf{GL}_3} % GL_3
 \newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
+\newcommand{\MatZ}{\mathbf{Mat}_3(\mathbb{Z})}        % Mat_3( ZZ )
 \newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
 \newcommand{\phip}{\phi_p}        % the golden ratio mod(p)
+\newcommand{\hence}{\Rightarrow}        % hence
 
@@
 \maketitle
+
+\tableofcontents
 
@@
 \begin{multline}
+	\label{computation_char_poly_of_a_tilde}
 \chi_{\tilde A}(\lambda) 
@@
 
+\subsection{Lucas primes give small degree $\log(p)$}
+
 \begin{Prop}
@@
 
+\subsection{Bounds on the degree $log(p) \prec \deg \prec p$}
+Motivated by geometric considerations, we call the degree $\deg$ the index $[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$.
+
+\begin{Prop}
+We have $O(log(p)) \le \deg \le O(p)$.
+\end{Prop}
+
+\begin{proof}
+The proof of the previous Theorem shows that $Z_{\SL}(\tilde A)$ is generated by  $\alpha^3(2 \alpha + 1)^{-1}, \alpha^{\deg}$.
+
+So, the goal is to show that the multiplicative order of $\phi_p$ is bounded by $\log(p)$ from below and by $p$ from above.
+
+First let us show $\log(p) < \deg$.
+Indeed, lifting $\phi_p$ to $\phi$ we notice that 
+\[
+\phi_p^k = 1 \hence (- \phi_p^{-1} )^{k} = (-1)^k \hence \phi^k + (- \phi^{-1} )^{k} - (1 + (-1)^k) \in p \mathbb{Z}
+.\] 
+Thus the degree $\deg > \log(p) - \epsilon$ for $p$ big enough and some small $\epsilon$.
+
+Finally, we shall show $\deg \le O(p)$.
+What we show in reality is that if $\sqrt 5 \in \mathbb F_p$, the degree $\deg$ divides $p-1$, $\deg | p-1$ and if $\sqrt 5 \notin \mathbb F_p$, $\deg | 2 p + 2$.
+
+Indeed, the first part is clear by Fermat's little theorem.
+To show the second, let's decompose 
+\[
+x^2 - x - 1 = (x - \alpha_1)(x - \alpha_2)
+,\] 
+where $\alpha_1, \alpha_2 \in \mathbb F_{p^2}$.
+The Frobenius automorphism $\mathrm{Frob}: t \mapsto t^p$ acts on the roots $\alpha_1, \alpha_2$ by permuting them; thus we have $\alpha_1^p = \alpha_2 = -1/\alpha_1$ implying $\alpha_1^{2p+2} = (-1)^2 = 1$.
+\end{proof}
+
+\section{Enter family}
+Let $f_{a,b}(t) = t^3 - a p t^2 - (b p + a) t - b$.
+Here $a, b \in \mathbb{Z} $.
+Note that $f_{2,1}(t) = f(t)$ is the characteristic polynomial of $A$.
+
+We can now define
+\begin{equation}
+\label{matrix_small_family}
+A_{a, b} = 
+\begin{pmatrix}
+0 & 0 & b \\
+1 & 0 & bp+a \\
+0 & 1 & ap
+\end{pmatrix}
+\in \MatZ,
+\end{equation}
+\begin{equation}
+\label{matrix_big_family}
+\tilde A_{a, b} = \Id + p A_{a, b} =
+\begin{pmatrix}
+1 & 0 & b p \\
+p & 1 & b p^2 + a p \\
+0 & p & 1 + a p^2
+\end{pmatrix}
+\in \SLp.
+\end{equation}
+Here $\tilde A_{a, b}$ is indeed invertible thanks to 
+\begin{multline}
+	\label{computation_char_poly_of_a_tilde_family}
+	\det({{\tilde{A}}_{a,b}})
+= p^3 \det \left(\frac{1}p \Id + A_{a,b}\right) 
+= - p^3 \det \left(\frac{-1}p \Id - A_{a,b}\right) \\
+= - p^3 f_{a,b}\left(\frac{-1}p\right)
+= 1 + a p^2 - (b p + a) p^2 + b p^3
+= 1
+.\end{multline}
+Actually, almost all of the matrices from $\SLp$ are similar to $\tilde A_{p, q}$.
+\begin{Prop}
+Any matrix $B \in \SLp$ whose minimal polynomial is its characteristic polynomial satisfies
+\[
+B = C \tilde A_{a, b} C^{-1} 
+,\] for some $a, b \in \mathbb{Z}, C \in \GL( \mathbb{Q} )$.
+\end{Prop}
+\begin{proof}
+Pick $C \in \GL( \mathbb{Q} ) $ such that 
+$$
+C^{-1} \frac{1}{p}(B - \Id) C =
+\begin{pmatrix}
+0 & 0 & r_0 \\
+1 & 0 & r_1 \\
+0 & 1 & r_2
+\end{pmatrix} \in 
+\mathbf{SL}_3(\mathbb{Q})
+.$$ 
+This is possible to do because of the condition on $B$ and the theory of Frobenius normal form over $ \mathbb{Q}$ from linear algebra.
+
+This matrix is actually integer $ C^{-1} \frac{1}{p}(B - \Id) C \in \MatZ$.
+Indeed, the coefficients in the last column are the coefficients of the characteristic polynomial. 
+Therefore, $ C^{-1} B C \in \SLp $.
+
+It is left to show that 
+$ C^{-1} B C = \tilde A_{a, b}$.
+This follows from the equation $\det C^{-1} B C = 1$ which reads
+\[
+1 + r_2 p - r_1 p^2 + r_0 p^3 = 1
+,\] 
+which implies
+ \[
+r_0 = b, r_1 = bp + a, r_2 = ap,
+\] 
+for some $a, b \in \mathbb{Z}$.
+\end{proof}
+
+\section{The Unknown}
+\begin{Prob}
+What is the behavior of $\Reg, \deg $ for an arbitrary member of the family $\tilde A_{a, b}$?
+\end{Prob}
+Currently I don't really understand how to treat ramification.
+
 \bibliography{mybib}

### 2024-05-27 draft_systoles.tex
family is everying

@@
 \newcommand{\GL}{\mathbf{GL}_3} % GL_3
+\newcommand{\PGL}{\mathbf{GL}_3} % PGL_3
 \newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
@@
 \newcommand{\hence}{\Rightarrow}        % hence
+\newcommand{\minpoly}{\mathrm{minpoly}}        % minimal polynomial
 
@@
 .\end{multline}
-Actually, almost all of the matrices from $\SLp$ are similar to $\tilde A_{p, q}$.
+Actually, a generic element in $\SLp$ is conjugate to $\tilde A_{a, b}$.
 \begin{Prop}
-Any matrix $B \in \SLp$ whose minimal polynomial is its characteristic polynomial satisfies
+An element $\tilde B \in \SLp$ is either
+\begin{enumerate}
+	\item $\Id$;
+	\item conjugate to 
 		\[
-B = C \tilde A_{a, b} C^{-1} 
-,\] for some $a, b \in \mathbb{Z}, C \in \GL( \mathbb{Q} )$.
+		\begin{pmatrix}
+		1 & p & 0 \\
+		0 & 1 & 0 \\
+		0 & 0 & 1
+		\end{pmatrix}
+		\] 
+		over $\mathbb{Q}$;
+	\item conjugate to $\tilde A_{a, b}$ over $\mathbb{Q}$ for some $a, b \in \mathbb{Z}$.
+\end{enumerate}
+% That is, for any $B \neq \Id \in \SLp$ we have
+% \[
+% B = C \tilde A_{a, b} C^{-1} 
+% ,\] for some $a, b \in \mathbb{Z}, C \in \GL( \mathbb{Q} )$.
+% Moreover, $a, b$ are uniquely defined and $C$ is unique as an element of $\PGL( \mathbb{Q} )$.
 \end{Prop}
+The proof goes by applying the Frobenius normal form over $ \mathbb{Q}$.
 \begin{proof}
+Define $B = \frac{1}{p}(\tilde B - \Id) \in \MatZ$.
+
+There are three cases:
+the minimal polynomial $\minpoly_B$ of $B$ is of degree $1, 2$ or $3$.
+\begin{enumerate}
+	\item $\deg \minpoly_B = 1$.  Hence $\tilde B$ is scalar, hence $\Id$;
+	\item $\deg \minpoly_B = 2$.  Hence $ \frac{\chi_B}{\minpoly_B}$ is linear, thus dividing $\minpoly_B$; we obtain $\chi_B$ decomposes into a product of linear multiples over $ \mathbb{Z}$
+	$$
+	\chi_{B} (\lambda) = (\lambda - a)^2 (\lambda - b), a, b \in \mathbb{Z}.
+	$$
+	The condition $ - p^3 \chi_{B}( \frac{-1}{p}) = \chi_{\tilde B}(0) = 1$ now reads
+	\[
+	(1 + a p)^2 (1 + b p ) = 1
+	,\] 
+	thus $a, b = 0$ if $p > 2$ and $B$ is nilpotent.
+	Therefore $\tilde B$ is conjugate to 
+	\[
+	\begin{pmatrix}
+	1 & p & 0 \\
+	0 & 1 & 0 \\
+	0 & 0 & 1
+	\end{pmatrix}
+	.\] 
+	\item
+	The Frobenius normal form of 
+	$B$ 
+	% $\frac{1}{p}(B - \Id)$ 
+	consists of the unique $3 \times 3$ block.
+
 	Pick $C \in \GL( \mathbb{Q} ) $ such that 
 	$$
-C^{-1} \frac{1}{p}(B - \Id) C =
+	C^{-1} B C =
 	\begin{pmatrix}
@@
 	.$$ 
-This is possible to do because of the condition on $B$ and the theory of Frobenius normal form over $ \mathbb{Q}$ from linear algebra.
+	This matrix is actually integer $ C^{-1} B C \in \MatZ$.
+	Indeed, the coefficients in the last column are the coefficients of the characteristic polynomial. 
+	Therefore, $ C^{-1} \tilde B C \in \SLp $.
 
-This matrix is actually integer $ C^{-1} \frac{1}{p}(B - \Id) C \in \MatZ$.
-Indeed, the coefficients in the last column are the coefficients of the characteristic polynomial. 
-Therefore, $ C^{-1} B C \in \SLp $.
-
-It is left to show that 
-$ C^{-1} B C = \tilde A_{a, b}$.
-This follows from the equation $\det C^{-1} B C = 1$ which reads
+	We now show that 
+	$ C^{-1} \tilde B C = \tilde A_{a, b}$.
+	This follows from the equation $\det (C^{-1} \tilde B C) = 1$ which reads
 	\[
@@
 	,\] 
-which implies
+	implying
 	 \[
@@
 	for some $a, b \in \mathbb{Z}$.
+\end{enumerate}
+
 \end{proof}
@@
 \end{Prob}
-Currently I don't really understand how to treat ramification.
 

### 2024-06-27 draft_systoles.tex
add todo till next time
The new full version

@@
 \newcommand{\red}{\textcolor{red}} % support of \red text
+
+\usepackage{listings}
+
+% Define SageMath style for highlighting
+\lstdefinestyle{sagemath}{
+    language=Python, % SageMath is based on Python
+    basicstyle=\ttfamily\footnotesize,
+    commentstyle=\color{gray},
+    keywordstyle=\color{blue},
+    stringstyle=\color{red},
+    frame=single,
+    breaklines=true,
+    showstringspaces=false,
+    tabsize=4,
+    captionpos=b,
+    morekeywords={R, PolynomialRing, QQ} % Add SageMath specific keywords here
+}
 
@@
 The centralizer of $\tilde A$ in $\SL$
-is generated by $A, A+\Id$.
+is generated by $A, 2 A+\Id$.
 $$Z_{\SL}(\tilde A) = \{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
@@
 
+TODO:
+* Study the problem for fixed p and changing $a, b$;
+here $a, b$ are such that $\Reg f_{a,b} < p/5$?
+
+\[
+\disc f_{a,b} 
+= a^2*p^4*b^2 - 2*a^3*p^3*b + a^4*p^2 + 4*p^3*b^3 - 6*a*p^2*b^2 - 6*a^2*p*b + 4*a^3 - 27*b^2
+.\]
+
+\section{A Strange example}
+Let $p$ a prime number with $p \equiv 1 \pmod 3$.
+Define
+\begin{equation}
+\label{matrix_small}
+A = 
+\begin{pmatrix}
+0 & 0 & 1 \\
+1 & 0 & 2p \\
+0 & 1 & p^2
+\end{pmatrix}
+\in \SL
+\end{equation}
+\begin{equation}
+\label{matrix_big}
+\tilde A = \Id + p A =
+\begin{pmatrix}
+1 & 0 & p \\
+p & 1 & 2 p^2 \\
+0 & p & 1 + p^3
+\end{pmatrix}
+\in \SLp
+\end{equation}
+The characteristic poly of $A$ is $ f(t) \assign \chi_A(t) = t^3 - p^2 t^2 - 2 p t - 1$.
+
+Let $\Q(\alpha)$ be the number field obtain by attaching the root of $f(t)$ to $\Q$.
+\begin{Prop}
+$\Q(\alpha) / \Q$ is a complex number field of degree $3$.
+\end{Prop}
+\red{This is terrible as the unit group then is of rank one.}
+
+\begin{proof}
+The discriminant of $f(t)$ equals
+$$\disc_f(p) = -4 p^3 - 27 .$$
+The discriminant is always negative $\disc_f(p) < 0$, thus $f$ has only one real root and the other roots are complex and conjugate to each other.
+\end{proof}
+
+\begin{Prop}
+$\mathbb{Z}[\alpha]^{*} = \{ \pm \alpha^a (1 + p \alpha)^b \mid a, b \in \mathbb{Z} \} $.
+\end{Prop}
+
+Note that $\alpha, 1 + p \alpha$ are not necessarily fundamental units in $\Q(\alpha)/\Q$ as we don't claim that the ring of integers of $\Q(\alpha) / \Q$ coincides with $\mathbb{Z}[\alpha]$.
+
+Note that $f(\alpha) = 0$ implies 
+\begin{gather}
+\alpha (\alpha ^2 - p^2 \alpha - 2 p) = 1
+\\
+(1 + p \alpha)^2 = \alpha^3
+\end{gather}
+It shows that $\alpha, 1 + p \alpha$ are indeed units.
+
+Let $\alpha_1 \in \R, \alpha_2 = \bar \alpha_3 \in \C \setminus \R$ be the three different roots of $f$.
+We shall compute them approximately.
+\begin{gather*}
+\alpha_1 = p^2 + \frac{2}{p} + O\left(\frac{1}{p}\right), \\
+\alpha_2 = - p^{-1} + i~p^{-\frac{5}{2}} + \frac{3}{2}p^{-4} + O\left(\frac{1}{p^4}\right), \\
+\alpha_3 = - p^{-1} - i~p^{-\frac{5}{2}} + \frac{3}{2}p^{-4} + O\left(\frac{1}{p^4}\right)
+.\end{gather*}
+
+\begin{Rem}
+A computation shows that for $p = 1000$ we have
+\begin{gather*}
+\alpha_1 = 1.0000000020000000\mathrm{e}{+6} , \\
+\alpha_2 = -0.000999999998500000 - i~ 3.162277652\mathrm{e}{-8}, \\
+\alpha_3 = -0.000999999998500000 + i~ 3.162277652\mathrm{e}{-8}
+.\end{gather*}
+It is not important that $p$ is not a prime in this case as the estimate works for any sufficiently large $p$.
+\end{Rem}
+
+\[
+f(t) = t^3 - a p^2 t^2 - (a + 1)p t - 1
+.\] 
+\[
+1 = (t^2 - a p^2 t - (a + 1)p) t, 
+t^3 =  (p t + 1) (a p t + 1)
+.\]
+
+\section{A Strange example v2}
+Let $p$ a prime number with $p \equiv 1 \pmod 3$.
+Define
+\begin{equation}
+A = 
+\begin{pmatrix}
+0 & 0 & 1 \\
+1 & 0 & 0 \\
+0 & 1 & -p^2
+\end{pmatrix}
+\in \SL
+\end{equation}
+\begin{equation}
+\tilde A = \Id + p A =
+\begin{pmatrix}
+1 & 0 & p \\
+p & 1 & 0 \\
+0 & p & 1 - p^3
+\end{pmatrix}
+\in \SLp
+\end{equation}
+The characteristic poly of $A$ is $ f(t) \assign \chi_A(t) = t^3 + p^2 t^2 - 1$.
+
+Let $\Q(\alpha)$ be the number field obtain by attaching the root of $f(t)$ to $\Q$.
+\begin{Prop}
+$\Q(\alpha) / \Q$ is a totally real number field of degree $3$.
+\end{Prop}
+
+\begin{proof}
+The discriminant of $f(t)$ equals
+$$\disc_f(p) = 4 p^6 - 27 .$$
+
+The discriminant is always positive $\disc_f(p) > 0$, thus the number field is totally real.
+\end{proof}
+
+\begin{Prop}
+$\mathbb{Z}[\alpha]^{*} = \{ \pm \alpha^a (1 + p \alpha)^b \mid a, b \in \mathbb{Z} \} $.
+\end{Prop}
+
+Note that $\alpha, 1 + p \alpha$ are not necessarily fundamental units in $\Q(\alpha)/\Q$ as we don't claim that the ring of integers of $\Q(\alpha) / \Q$ coincides with $\mathbb{Z}[\alpha]$.
+
+\begin{proof}
+We have
+\begin{gather}
+\alpha (\alpha ^2 + p^2 \alpha) = 1
+\\
+(1 - p \alpha)(1 + p \alpha) = \alpha^3
+\end{gather}
+It shows that $\alpha, 1 + p \alpha$ are indeed units.
+
+Let $\alpha_1 < \alpha_2 < \alpha_3 $ be the three different roots of $f$.
+We shall compute them approximately.
+\begin{gather*}
+\alpha_1 = -p^2 + O\left(p^{-4} \right), \\
+\alpha_2 = - p^{-1}   + O\left(p^{-4} \right), \\
+\alpha_3 = p^{-1} + O\left(p^{-4} \right)
+.\end{gather*}
+
+\begin{Rem}
+A computation shows that for $p = 100$ we have
+\begin{gather*}
+\alpha_1 = -9999.99999999000 , \\
+\alpha_2 = -0.01000000500000625, \\
+\alpha_3 = 0.00999999500000625
+.\end{gather*}
+It is not important that $p$ is not a prime in this case as the estimate works for any sufficiently large $p$.
+\end{Rem}
+
+By the definition of the regulator we have
+\begin{multline}
+\Reg(\alpha, p\alpha + 1) = 
+\begin{vmatrix}
+\ln |\alpha_1| 		& \ln |\alpha_3| \\
+\ln |p \alpha_1 + 1| 	& \ln |p \alpha_3 + 1|
+\end{vmatrix} 
+\\
+\approx
+\begin{vmatrix}
+\ln |-p^2| 	& \ln |\frac{1}{p}| \\
+\ln |-p^3 + 1| 	& \ln |2|
+\end{vmatrix}
+=
+\ln(p^2)
+\ln 2
++
+\ln(p^3 - 1) \ln(p)
+\\
+\approx \ln(p) ( 3 \ln(p) + 2 \ln 2 )
+.\end{multline}
+Therefore $ \Reg(\alpha, p\alpha + 1) > 0 $ for all prime $p$.
+
+Hence $\alpha, p\alpha + 1$ are independent units.
+
+\begin{Rem}
+For example, for $p = 547$ SageMath computes the regulator to be approximately $\Reg = 127.978045931846$.
+
+\begin{lstlisting}[style=sagemath, caption=SageMath code example]
+p = Primes().unrank(100) # p = 547
+R.<x> = PolynomialRing(QQ)
+P = x^3 + p^2 * x^2  - 1
+if not P.is_irreducible():
+    print("NOT IRREDUCIBLE")
+    break
+K.<a> = QQ.extension(P)
+print(K.regulator())
+\end{lstlisting}
+Whereas the above estimate gives $127.97804593184651$.
+
+These numbers coincide up to $10^{-12} $.
+\end{Rem}
+
+It is left to prove that they are fundamental units in $\mathbb{Z}[\alpha]$.
+By Corollary 5.9 from Conrad it is sufficient to check
+\[
+\frac{16 \Reg(\alpha, p\alpha + 1) }
+{(\ln(\disc_f/4))^2} 
+< 2
+.\] 
+Substituting, we obtain
+\[
+\frac{16 \Reg(\alpha, 2\alpha + 1) }
+{(\ln(\disc_f/4))^2} 
+\approx
+\frac{16 
+\ln(p) ( 3 \ln(p) + 2 \ln 2 )
+ }
+{(\ln(
+(4 p^6 - 27)
+/4))^2} 
+.\] 
+Asymptotically, the latter equals
+\[
+\stackrel{p \to \infty}{\longrightarrow}
+\frac{48
+\ln(p)^2
+}
+{(\ln(
+p^6 ))^2} = \frac{4}{3}
+.\] 
+Therefore it is $< 2$ for big enough $p$, QED.
+\end{proof}
+
+\begin{Rem}
+We just proved that the regulator is approximately 
+\[
+\ln(p) ( 3 \ln(p) + 2 \ln 2 )
+,\] 
+which is close to $\ln^2 p$ we wanted from the beginning.
+\end{Rem}
+
+\section{Centralizers}
+\begin{Prop}
+The centralizer of $\tilde A$ in $\SL$
+is generated by $A, p A+\Id$.
+$$Z_{\SL}(\tilde A) = \{ A^a (p A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
+\end{Prop}
+\begin{proof}
+Since $\tilde A$ is regular, its centralizer in $ \Mat $ is $\C \left< \Id, A, A^2 \right> $.
+Now, 
+\[
+\C \left< \Id, A, A^2 \right> \cap \SL 
+\subset \mathbb{Z} 
+\left< \Id, A, A^2 \right>
+.\] 
+Indeed,
+\begin{equation}
+\Id = 
+\begin{pmatrix}
+1 & 0 & 0 \\
+0 & 1 & 0 \\
+0 & 0 & 1
+\end{pmatrix}, 
+A = 
+\begin{pmatrix}
+0 & 0 & 1 \\
+1 & 0 & 0 \\
+0 & 1 & -p^2
+\end{pmatrix}, 
+A^2 = 
+\begin{pmatrix}
+0 & 1 & -p^2 \\
+0 & 0 & 1 \\
+1 & -p^2 & p^4
+\end{pmatrix}, 
+\end{equation}
+Considering the first matrix column we see that if a complex combination has integer coefficients, it is in fact integer combination.
+
+Moreover, 
+the centralizer of $\tilde A$ is a group, therefore it lies inside the multiplicative group of $\mathbb{Z}[A]$
+\[
+Z_{\SL}(\tilde A)  \subset 
+\C \left< \Id, A, A^2 \right> \cap \SL 
+\subset \mathbb{Z}[A]^*
+.\] 
+There is an isomorphism of $\mathbb{Z}$-algebras $\mathbb{Z}[A] \simeq \mathbb{Z}[x]/(f(x)) = \mathbb{Z}[\alpha]$.
+Applying Proposition \ref{prop_multiplicative_group_structure} end the proof
+\[
+Z_{\SL}(\tilde A) 
+\subset \mathbb{Z}[A]^* =\{ \pm A^a (p A + \Id)^b \mid a, b \in \mathbb{Z} \}.\] 
+\end{proof}
+
+We are to study the centralizer of $\tilde A$ in $\SLp$.
+\[
+Z_{\SLp}(\tilde A) \subset 
+Z_{\SL}(\tilde A) \cong \mathbb{Z}^2
+.\]
+
+\begin{Prop}
+	The index of the one centralizer inside the other \emph{does not depend on $p$} and equals $3$.
+\end{Prop}
+\begin{proof}
+As $A$ is equivalent to the cyclic permutation matrix $\pmod p$ it follows $A^3 \in \SLp$.
+
+Thus the smaller centralizer is generated $A^3, \tilde A$
+\[
+Z_{\SLp}(\tilde A)
+= \{ A^{3a} (p A + \Id)^b \mid a, b \in \mathbb{Z} \}
+,\] 
+and it's clear that the index is $3$.
+\end{proof}
+
 \bibliography{mybib}

### 2024-06-27 draft_systoles.tex
Removed Lucas and family, add code for small p

--- 2024-05-26 draft_systoles.tex
@@
 
+\usepackage{array}
+
 \usepackage{xcolor} % support of \red text
 \newcommand{\red}{\textcolor{red}} % support of \red text
+
+\usepackage{listings}
+
+% Define SageMath style for highlighting
+\lstdefinestyle{sagemath}{
+    language=Python, % SageMath is based on Python
+    basicstyle=\ttfamily\footnotesize,
+    commentstyle=\color{gray},
+    keywordstyle=\color{blue},
+    stringstyle=\color{red},
+    frame=single,
+    breaklines=true,
+    showstringspaces=false,
+    tabsize=4,
+    captionpos=b,
+    morekeywords={R, PolynomialRing, QQ} % Add SageMath specific keywords here
+}
 
@@
 \begin{itemize}
-\item $p$ a prime number.
+\item $p > 2$ a prime number.
 \item $\SL$ the special linear group over $\mathbb{Z}$.
@@
 0 & 0 & 1 \\
-1 & 0 & p+2 \\
-0 & 1 & 2p
+1 & 0 & 0 \\
+0 & 1 & -p^2
 \end{pmatrix}
@@
 1 & 0 & p \\
-p & 1 & 2 p + p^2 \\
-0 & p & 1 + 2 p^2
+p & 1 & 0 \\
+0 & p & 1 - p^3
 \end{pmatrix}
@@
 \item If $C$ is a matrix, $\chi_C(\lambda) \assign \det(\lambda \Id - C)$ is the characteristics polynomial.
-\item $ f(t) \assign \chi_A(t) = t^3 - 2p t^2 - (p + 2) t - 1$.
+\item $ f(t) \assign \chi_A(t) = t^3 + p^2 t^2 - 1$.
 
@@
 \section{Intro}
-Recall the Lucas primes (see \url{https://en.wikipedia.org/wiki/Lucas_number#Lucas_primes}, \url{https://t5k.org/top20/page.php?id=48})
-\begin{equation}
-	\label{the_primes_sequence}
-2, 3, 7, 11, 29, 47, 199, 521, 2207, 3571, 9349, 3010349, 54018521, 370248451, 6643838879, \ldots
-\end{equation}
-In this notes we establish the following results:
+In this note we establish the following results:
 \begin{enumerate}
@@
 The regularor of $\tilde A$
-grows as $\approx \ln^2(p)$.
+grows as $\approx 3 \ln^2(p)$.
 
-That is, $\Reg ( \Q(\alpha)/\Q ) \approx \ln^2(p) $ for $\alpha$ a root of $\chi_{\tilde A}$.
+That is, $\Reg ( \Q(\alpha)/\Q ) \approx 3 \ln^2(p) $ for $\alpha$ a root of $\chi_{\tilde A}$.
 
@@
 The index of the centralizers
-$ [Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A)] $
-grows as $O(\ln p)$ for $p$ a Lucas prime.
+$ [Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A)] = 3$
+does not depend on $p$.
 \end{enumerate}
 This might be interesting.
-\section{Regulator}
-We refer to Keith Conrad's write-up on Dirichlet's unit theorem and regulators \cite{conraddirichlet} for the definitions.
-The current proof mimics the proof of Theorem 5.12 of Conrad.
 
-Notice that 
-\begin{multline}
-	\label{computation_char_poly_of_a_tilde}
-\chi_{\tilde A}(\lambda) 
-= \det(\lambda \Id - \tilde A) \\
-= \det(\lambda \Id - (\Id + p A))
-= \det((\lambda - 1) \Id - p A)\\
-= p^3 \det(\frac{\lambda - 1}p \Id - A)
-= p^3 \chi_A(\frac{\lambda - 1}p)
-.\end{multline}
-Hence adding the root of $\chi_{A}$ or $\chi_{\tilde A}$ result in the same field; therefore we reduce to showing that 
-$\Reg ( \Q(\alpha)/\Q ) \approx \ln^2(p) $ for $\alpha$ a root of $\chi_{A}(t) = f(t) = t^3 - (2p t^2 + (p + 2) t + 1) $.
+\section{Regulators}
+We mimic the proofs from a Keigh Conrad's write-up on the Dirichlet unit theorem \cite{conraddirichlet}.
 
-\begin{Lem}
-$\Q(\alpha)/\Q$ is totally real of degree $3$ for primes $p \ne 2$.
-\end{Lem}
+Let $\Q(\alpha)$ be the number field obtain by attaching the root of $f(t)$ to $\Q$.
+\begin{Prop}
+$\Q(\alpha) / \Q$ is a totally real number field of degree $3$.
+\end{Prop}
+
 \begin{proof}
-$f(t)$ is irreducible over $\Q$; indeed, by the rational roots theorem it's sufficient to check $\pm 1$:
-$$f(1) = - 3 p - 2, f(-1) = - p .$$
+First, $f(t)$ is irreducible as it has no rational roots: $f(1) = p^2 \ne 0, f(-1) = p^2 - 2 \ne 0 $.
 
-A simple computation shows that the discriminant of $f(t)$ equals 
-\[
-	\disc_f(p) = 4 p^4 - 12 p^3 + 4 p^2 - 24 p + 5
-.\]
-If $p > 4$ the discriminant  $\disc_f(p) > 0$ is positive.
-Therefore the cubic extension is totally real.
+The discriminant of $f(t)$ equals
+$$\disc_f(p) = 4 p^6 - 27 .$$
+
+The discriminant is always positive $\disc_f(p) > 0$, thus the number field is totally real.
 \end{proof}
@@
 \label{prop_multiplicative_group_structure}
-$\mathbb{Z}[\alpha]^{*} = \{ \pm \alpha^a (2\alpha + 1)^b \mid a, b \in \mathbb{Z} \} $.
+$\mathbb{Z}[\alpha]^{*} = \{ \pm \alpha^a (1 + p \alpha)^b \mid a, b \in \mathbb{Z} \} $.
 \end{Prop}
-Note that $\alpha, 2\alpha + 1$ are not necessarily fundamental units in $\Q(\alpha)/\Q$ as we don't claim that the ring of integers of $\Q(\alpha) / \Q$ coincides with $\mathbb{Z}[\alpha]$.
+
+Note that $\alpha, 1 + p \alpha$ are not necessarily fundamental units in $\Q(\alpha)/\Q$ as we don't claim that the ring of integers of $\Q(\alpha) / \Q$ coincides with $\mathbb{Z}[\alpha]$.
+
 \begin{proof}
-Note that $f(\alpha) = 0$ implies 
+We have
 \begin{gather}
-\alpha (\alpha^2 - 2p \alpha - (p + 2)) = 1
+\alpha (\alpha ^2 + p^2 \alpha) = 1
 \\
-\label{equation_units}
-(1 + 2 \alpha) (1 + p \alpha) = \alpha^3
+(1 - p \alpha)(1 + p \alpha) = \alpha^3
 \end{gather}
-It shows that $\alpha, 1 + 2 \alpha$ are indeed units.
+It shows that $\alpha, 1 + p \alpha$ are indeed units.
 
-Let $\alpha_1 > \alpha_2 > \alpha_3 \in \R$ be the three different roots of $f$.
+Let $\alpha_1 < \alpha_2 < \alpha_3 $ be the three different roots of $f$.
 We shall compute them approximately.
 \begin{gather*}
-\alpha_1 = 2 p + \frac{1}{2} + O\left(\frac{1}{p}\right), \\
-\alpha_2 = - \frac{1}{ p } + O\left(\frac{1}{p^4}\right), \\
-\alpha_3 = - \frac{1}{2} + O\left(\frac{1}{p}\right)
+\alpha_1 = -p^2 + O\left(p^{-4} \right), \\
+\alpha_2 = - p^{-1}   + O\left(p^{-4} \right), \\
+\alpha_3 = p^{-1} + O\left(p^{-4} \right)
 .\end{gather*}
-% from sympy import *
-% [ins] In [2]: x = symbols('x')
-% [ins] In [3]: p = 10000
-% [ins] In [12]: f = x**3 - 2 * p * x**2 - (p + 2) * x - 1
-% [ins] In [13]: f = Poly(f)
-% [ins] In [16]: nroots(f)
-% Out[16]: [-0.499987498124648, -0.000100000000000100, 20000.5000874981]
+
 \begin{Rem}
-
-A computation shows that for $p = 10000$
-we have
- \[
-\alpha_1 = 20000.5000874981 ,
-\alpha_2 = -0.000100000000000100 ,
-\alpha_3 = -0.499987498124648 
-.\] 
+A computation shows that for $p = 100$ we have
+\begin{gather*}
+\alpha_1 = -9999.99999999000 , \\
+\alpha_2 = -0.01000000500000625, \\
+\alpha_3 = 0.00999999500000625
+.\end{gather*}
 It is not important that $p$ is not a prime in this case as the estimate works for any sufficiently large $p$.
@@
 \begin{multline}
-\Reg(\alpha, 2\alpha + 1) = 
+\Reg(\alpha, p\alpha + 1) = 
 \begin{vmatrix}
-\ln |\alpha_1| 		& \ln |\alpha_2| \\
-\ln |2 \alpha_1 + 1| 	& \ln |2 \alpha_2 + 1|
+\ln |\alpha_1| 		& \ln |\alpha_3| \\
+\ln |p \alpha_1 + 1| 	& \ln |p \alpha_3 + 1|
 \end{vmatrix} 
@@
 \begin{vmatrix}
-\ln |2p + \frac{1}{2}| 	& \ln |\frac{-1}{p}| \\
-\ln |4 p + 2| 	& \ln |\frac{-2}{p} + 1|
+\ln |-p^2| 	& \ln |\frac{1}{p}| \\
+\ln |-p^3 + 1| 	& \ln |2|
 \end{vmatrix}
 =
-\ln(2p + \frac{1}{2})
-(\ln(p - 2) - \ln(p))
+\ln(p^2)
+\ln 2
 +
-\ln(4p + 2) \ln(p)
+\ln(p^3 - 1) \ln(p)
 \\
-=
-\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
+\approx \ln(p) ( 3 \ln(p) + 2 \ln 2 )
 .\end{multline}
-Therefore $ \Reg(\alpha, 2\alpha + 1) > 0 $ for all prime $p$.
+Therefore $ \Reg(\alpha, p\alpha + 1) > 0 $ for all prime $p$.
 
-Hence $\alpha, 2\alpha + 1$ are independent units.
+Hence $\alpha, p\alpha + 1$ are independent units.
+
+\begin{Rem}
+For example, for $p = 73$ SageMath computes the regulator to be approximately $\Reg = 61.1719663782187$.
+
+\begin{lstlisting}[style=sagemath, caption=SageMath code]
+p = Primes().unrank(20) # p = 73
+R.<x> = PolynomialRing(QQ)
+P = x^3 + p^2 * x^2  - 1
+K.<a> = QQ.extension(P)
+print(K.regulator())
+\end{lstlisting}
+Whereas the above estimate gives $61.1719663782957$.
+These numbers coincide up to $10^{-11} $.
+
+Another example, for $p = 547$ SageMath computes the regulator to be approximately $\Reg = 127.978045931846$.
+
+\begin{lstlisting}[style=sagemath, caption=SageMath code]
+p = Primes().unrank(100) # p = 547
+R.<x> = PolynomialRing(QQ)
+P = x^3 + p^2 * x^2  - 1
+K.<a> = QQ.extension(P)
+print(K.regulator())
+\end{lstlisting}
+Whereas the above estimate gives $127.97804593184651$.
+These numbers coincide up to $10^{-12} $.
+
+\end{Rem}
 
@@
 \[
-\frac{16 \Reg(\alpha, 2\alpha + 1) }
+\frac{16 \Reg(\alpha, p\alpha + 1) }
 {(\ln(\disc_f/4))^2} 
@@
 \approx
-\frac{16 (
-\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
-) }
+\frac{16 
+\ln(p) ( 3 \ln(p) + 2 \ln 2 )
+ }
 {(\ln(
-(p^4 + 2 p^3 - 5 p^2 - 6 p - 23)
+(4 p^6 - 27)
 /4))^2} 
@@
 \stackrel{p \to \infty}{\longrightarrow}
-\frac{16 
+\frac{48
 \ln(p)^2
@@
 {(\ln(
-p^4 ))^2} = 1
+p^6 ))^2} = \frac{4}{3}
 .\] 
@@
 \begin{Rem}
-One can do the estimate more carefully, but for now postpone it. 
+In fact, the function
+\[
+\frac{16 
+\ln(p) ( 3 \ln(p) + 2 \ln 2 )
+}
+{(\ln(
+(4 p^6 - 27)
+/4))^2} 
+\] 
+is smaller then $2$
+for all $p > 2.6$.
+It monotonously decreases to the asymptotic value $\frac{4}{3}$.
+
+The first several values of $g(p) \eqdef 
+\frac{16 \Reg(\alpha, 2\alpha + 1) }
+{(\ln(\disc_f/4))^2} 
+$ are presented in the following
+\begin{table}[h!]
+\centering
+\begin{tabular}{|c|c|}
+\hline
+$p$ & $g(p)$ \\
+\hline
+2 & 2.3005757200277737 \\
+3 & 0.948774985707828 \\
+5 & 1.7162543373046009 \\
+7 & 0.824986119068219 \\
+11 & 1.5902803808431747 \\
+13 & 1.573545207927126 \\
+17 & 1.5508005143276846 \\
+19 & 1.5425857167054255 \\
+23 & 1.529835319790773 \\
+29 & 1.5163082966042292 \\
+31 & 1.5127547443392193 \\
+37 & 1.5039633069049658 \\
+41 & 1.4992465878896728 \\
+43 & 1.4971456295557026 \\
+47 & 1.4933611793019335 \\
+53 & 1.4885186045113814 \\
+59 & 1.484436992267399 \\
+61 & 1.483211643289589 \\
+67 & 1.4798674282030506 \\
+71 & 1.4778740553564413 \\
+73 & 1.4769381932859227 \\
+79 & 1.4743421857047407 \\
+83 & 1.4727660220987417 \\
+89 & 1.4705979224908206 \\
+97 & 1.4680152465868979 \\
+101 & 1.4668359850881518 \\
+103 & 1.466271166102171 \\
+107 & 1.4651872592939545 \\
+109 & 1.4646667672702098 \\
+113 & 1.4636655275160453 \\
+127 & 1.4605230591156473 \\
+\hline
+\end{tabular}
+\caption{Values of $p$ and $g(p)$}
+\label{table:gp}
+\end{table}
+\newpage
+which we computed using SageMath
+\begin{lstlisting}[style=sagemath, caption=SageMath code]
+for i, p in enumerate(Primes()):
+    if i > 30:
+        break
+    R.<x> = PolynomialRing(QQ)
+    P = x^3 + p^2 * x^2 - 1
+    K.<a> = QQ.extension(P)
+    rg = K.regulator()
+    print(p, float(16 * rg * ln( p^6 - 27/4 )^(-2)))
+\end{lstlisting}
 \end{Rem}
+
 \begin{Rem}
@@
 \[
-\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
+\ln(p) ( 3 \ln(p) + 2 \ln 2 )
 ,\] 
-which is close to $\ln^2 p$ we wanted from the beginning.
+which is close to $3 \ln^2 p$ we wanted from the beginning.
 \end{Rem}
@@
 \begin{Prop}
-	\label{proposition_centralizer_in_sl}
 The centralizer of $\tilde A$ in $\SL$
-is generated by $A, A+\Id$.
-$$Z_{\SL}(\tilde A) = \{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
+is generated by $A, p A+\Id$.
+$$Z_{\SL}(\tilde A) = \{ A^a (p A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
 \end{Prop}
@@
 \subset \mathbb{Z} 
-\left< \Id, A, A^2 \right>
+\left< \Id, A, A^2 \right> = \mathbb{Z}[A]
 .\] 
@@
 0 & 0 & 1 \\
-1 & 0 & p+2 \\
-0 & 1 & 2p
+1 & 0 & 0 \\
+0 & 1 & -p^2
 \end{pmatrix}, 
@@
 \begin{pmatrix}
-0 & 1 & 2p \\
-0 & p+1 & 2p^2+4p+1 \\
-1 & p & 4p^2+p+2
+0 & 1 & -p^2 \\
+0 & 0 & 1 \\
+1 & -p^2 & p^4
 \end{pmatrix}, 
@@
 Z_{\SL}(\tilde A) 
-\subset \mathbb{Z}[A]^* =\{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \}.\] 
+\subset \mathbb{Z}[A]^* =\{ \pm A^a (p A + \Id)^b \mid a, b \in \mathbb{Z} \}.\] 
 \end{proof}
@@
 .\]
-In general this subgroup can be difficult to describe; 
-that leads to cosidering \emph{Lucas primes}.
-
-\subsection{Lucas primes give small degree $\log(p)$}
 
 \begin{Prop}
-Let $p$ be a Lucas prime.
-Let $k$ be the integer part of $\log_{\phi} p$ where $\phi$ is the golden ratio.
-
-The centralizer $Z_{\SLp}(\tilde A)$ contains $\tilde A$ and $A^{4k}$.
+	The index of the one centralizer inside the other \emph{does not depend on $p$} and equals $3$.
 \end{Prop}
 \begin{proof}
-The only thing to prove is that 
-$A^{4k}  \in \SLp$.
+As $A$ is equivalent to the cyclic permutation matrix $\pmod p$ it follows $A^3 \in \SLp$.
 
-It suffices to show that the eigenvalues of $A \pmod p$ are $4k$-th roots of unity.
-Computing
+Thus the smaller centralizer is generated $A^3, \tilde A$
 \[
-\chi_{A}(t) = f(t) \equiv t^3 - 2 t - 1 = (t + 1) (t^2 - t - 1) \pmod p
+Z_{\SLp}(\tilde A)
+= \{ A^{3a} (p A + \Id)^b \mid a, b \in \mathbb{Z} \}
 ,\] 
-shows that it is left to work with the golden ratio in $\mathbb F_p$ which we denote by $\phip$.
-That is, $\phip \in \mathbb F_{p^2}$ satisfies $\phip^2 - \phip - 1 = 0$.
-
-By the definition of a Lucas number we have 
-\begin{equation}
-	\label{eq_lucas_prime_definition}
-p = \phi^k + (-\phi)^{-k} 
-\end{equation}
-where $\phi$ is a root of $x^2 - x - 1$.
-
-The RHS of \eqref{eq_lucas_prime_definition} being invariant under the change $\phi \to (-\phi)^{-1}$ manifests it as a symmetric polynomial in the roots of $x^2 - x - 1$, thus having a presentation
-\[
-\phi^k + (-\phi)^{-k} = P( \phi, (-\phi)^{-1})
-,\] 
-where $P$ is a \emph{universal} polynomial.
-This observation justifies that the Equation \eqref{eq_lucas_prime_definition} can be taken modulo $p$ to have the form
-\[
-0 = \phip^k + (-\phip)^{-k} 
-,\] 
-which implies 
-\[
-1 = \phip^{4k} 
-.\] 
+and it's clear that the index is $3$.
 \end{proof}
-
-\begin{Thm}
-	The index of the centralizers is bounded by $4 \log_{\phi} p$
-\[
-	[Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A)] \le 4 \log_{\phi} p
-.\] 
-In particular, it grows as $O(\ln p)$.
-\end{Thm}
-\begin{proof}
-Identify $Z_{\SL}(\tilde A) \cong \mathbb{Z}/2\mathbb{Z} \oplus \mathbb{Z}^2$ as in Proposition \ref{proposition_centralizer_in_sl}.
-
-Observe that by \eqref{equation_units} we have
-\[
-1 + p \alpha = \alpha^3 (2 \alpha + 1)^{-1} 
-.\]
-
-By the previous Proposition 
-$Z_{\SLp}(\tilde A)$ contains $\binom{3}{-1}, \binom{4k}{0}$;
-Clearly, the index
-\[
-\mathbb{Z} \left< \binom{3}{-1}, \binom{4k}{0}\right> \subset \mathbb{Z}^2
-,\] 
-equals $4k \approx 4 \log_{\phi} p$ and the index 
-$ [Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$
-has to divide it.
-\end{proof}
-
-\subsection{Bounds on the degree $log(p) \prec \deg \prec p$}
-Motivated by geometric considerations, we call the degree $\deg$ the index $[Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$.
-
-\begin{Prop}
-We have $O(log(p)) \le \deg \le O(p)$.
-\end{Prop}
-
-\begin{proof}
-The proof of the previous Theorem shows that $Z_{\SL}(\tilde A)$ is generated by  $\alpha^3(2 \alpha + 1)^{-1}, \alpha^{\deg}$.
-
-So, the goal is to show that the multiplicative order of $\phi_p$ is bounded by $\log(p)$ from below and by $p$ from above.
-
-First let us show $\log(p) < \deg$.
-Indeed, lifting $\phi_p$ to $\phi$ we notice that 
-\[
-\phi_p^k = 1 \hence (- \phi_p^{-1} )^{k} = (-1)^k \hence \phi^k + (- \phi^{-1} )^{k} - (1 + (-1)^k) \in p \mathbb{Z}
-.\] 
-Thus the degree $\deg > \log(p) - \epsilon$ for $p$ big enough and some small $\epsilon$.
-
-Finally, we shall show $\deg \le O(p)$.
-What we show in reality is that if $\sqrt 5 \in \mathbb F_p$, the degree $\deg$ divides $p-1$, $\deg | p-1$ and if $\sqrt 5 \notin \mathbb F_p$, $\deg | 2 p + 2$.
-
-Indeed, the first part is clear by Fermat's little theorem.
-To show the second, let's decompose 
-\[
-x^2 - x - 1 = (x - \alpha_1)(x - \alpha_2)
-,\] 
-where $\alpha_1, \alpha_2 \in \mathbb F_{p^2}$.
-The Frobenius automorphism $\mathrm{Frob}: t \mapsto t^p$ acts on the roots $\alpha_1, \alpha_2$ by permuting them; thus we have $\alpha_1^p = \alpha_2 = -1/\alpha_1$ implying $\alpha_1^{2p+2} = (-1)^2 = 1$.
-\end{proof}
-
-\section{Enter family}
-Let $f_{a,b}(t) = t^3 - a p t^2 - (b p + a) t - b$.
-Here $a, b \in \mathbb{Z} $.
-Note that $f_{2,1}(t) = f(t)$ is the characteristic polynomial of $A$.
-
-We can now define
-\begin{equation}
-\label{matrix_small_family}
-A_{a, b} = 
-\begin{pmatrix}
-0 & 0 & b \\
-1 & 0 & bp+a \\
-0 & 1 & ap
-\end{pmatrix}
-\in \MatZ,
-\end{equation}
-\begin{equation}
-\label{matrix_big_family}
-\tilde A_{a, b} = \Id + p A_{a, b} =
-\begin{pmatrix}
-1 & 0 & b p \\
-p & 1 & b p^2 + a p \\
-0 & p & 1 + a p^2
-\end{pmatrix}
-\in \SLp.
-\end{equation}
-Here $\tilde A_{a, b}$ is indeed invertible thanks to 
-\begin{multline}
-	\label{computation_char_poly_of_a_tilde_family}
-	\det({{\tilde{A}}_{a,b}})
-= p^3 \det \left(\frac{1}p \Id + A_{a,b}\right) 
-= - p^3 \det \left(\frac{-1}p \Id - A_{a,b}\right) \\
-= - p^3 f_{a,b}\left(\frac{-1}p\right)
-= 1 + a p^2 - (b p + a) p^2 + b p^3
-= 1
-.\end{multline}
-Actually, almost all of the matrices from $\SLp$ are similar to $\tilde A_{p, q}$.
-\begin{Prop}
-Any matrix $B \in \SLp$ whose minimal polynomial is its characteristic polynomial satisfies
-\[
-B = C \tilde A_{a, b} C^{-1} 
-,\] for some $a, b \in \mathbb{Z}, C \in \GL( \mathbb{Q} )$.
-\end{Prop}
-\begin{proof}
-Pick $C \in \GL( \mathbb{Q} ) $ such that 
-$$
-C^{-1} \frac{1}{p}(B - \Id) C =
-\begin{pmatrix}
-0 & 0 & r_0 \\
-1 & 0 & r_1 \\
-0 & 1 & r_2
-\end{pmatrix} \in 
-\mathbf{SL}_3(\mathbb{Q})
-.$$ 
-This is possible to do because of the condition on $B$ and the theory of Frobenius normal form over $ \mathbb{Q}$ from linear algebra.
-
-This matrix is actually integer $ C^{-1} \frac{1}{p}(B - \Id) C \in \MatZ$.
-Indeed, the coefficients in the last column are the coefficients of the characteristic polynomial. 
-Therefore, $ C^{-1} B C \in \SLp $.
-
-It is left to show that 
-$ C^{-1} B C = \tilde A_{a, b}$.
-This follows from the equation $\det C^{-1} B C = 1$ which reads
-\[
-1 + r_2 p - r_1 p^2 + r_0 p^3 = 1
-,\] 
-which implies
- \[
-r_0 = b, r_1 = bp + a, r_2 = ap,
-\] 
-for some $a, b \in \mathbb{Z}$.
-\end{proof}
-
-\section{The Unknown}
-\begin{Prob}
-What is the behavior of $\Reg, \deg $ for an arbitrary member of the family $\tilde A_{a, b}$?
-\end{Prob}
-Currently I don't really understand how to treat ramification.
 

### 2024-06-27 draft_systoles.tex
Minor misprints correction

@@
 	The index of the one centralizer inside the other \emph{does not depend on $p$} and equals $3$.
+\[
+	[Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A) ] = 3
+.\] 
 \end{Prop}
 \begin{proof}
-As $A$ is equivalent to the cyclic permutation matrix $\pmod p$ it follows $A^3 \in \SLp$.
+As $A \pmod p$ is equivalent to the cyclic permutation matrix, it follows that $A^3 \in \SLp$.
 
-Thus the smaller centralizer is generated $A^3, \tilde A$
+Thus the smaller centralizer is generated by $A^3, \tilde A$
 \[
@@
 ,\] 
-and it's clear that the index is $3$.
+and it is clear that the index is $3$.
 \end{proof}

### 2024-07-10 draft_systoles.tex
add the SL_n case for n prime

@@
 \newcommand{\Mat}{\mathbf{Mat}_3(\mathbb{C})}        % Mat_3( C )
+\newcommand{\Matn}{\mathbf{Mat}_n(\mathbb{C})}        % Mat_n( C )
 \newcommand{\MatZ}{\mathbf{Mat}_3(\mathbb{Z})}        % Mat_3( ZZ )
+\newcommand{\MatnZ}{\mathbf{Mat}_n(\mathbb{Z})}        % Mat_n( ZZ )
 \newcommand{\SLp}{\Gamma_p}        % a congruence subgroup
@@
 \newcommand{\minpoly}{\mathrm{minpoly}}        % minimal polynomial
+
+\newcommand{\SLn}{\mathbf{SL}_n(\mathbb{Z})}        % SL_n(Z)
 
@@
 
+\section{From $\SL$ to $\SLn$}
+
+Consider
+\begin{itemize}
+\item $p > 2$ a prime number;
+\item $\SLn$ the $n$-th special linear group over $\mathbb{Z}$;
+\item $\SLp$ the $p$th congruence subgroup of $\SLn$;
+\item $Z_G(x)$ the centralizer of $x \in G$;
+\item $ f(t) \eqdef t^n - (1 + pt) (1 + 2pt) \ldots (1 + (n - 1) p t)$;
+\item If $C$ is a matrix, $\chi_C(\lambda) \assign \det(\lambda \Id - C)$ is the characteristics polynomial.
+\item Let $A$ be the irreducible operator with $\chi_A(t) = f(t)$;
+
+For example, when $n = 4$ $A$ takes a form
+\begin{equation}
+A = 
+\begin{pmatrix}
+0 & 0 & 0 & 1 \\
+1 & 0 & 0 & 6 p \\
+0 & 1 & 0 & 11 p^2 \\
+0 & 0 & 1 & 6 p^3
+\end{pmatrix}
+\in {\mathbf{SL}_4(\mathbb{Z})};        % SL_3(Z)
+\end{equation}
+\item 
+Define $\tilde A \eqdef \Id + p A$.
+
+For example, when $n = 4$ $\tilde A$ takes a form
+\begin{equation}
+\tilde A = \Id + p A =
+\begin{pmatrix}
+1 & 0 & 0 & p \\
+p & 1 & 0 & 6 p^2 \\
+0 & p & 1 & 11 p^3 \\
+0 & 0 & p & 1 + 6 p^4
+\end{pmatrix}
+\in \SLp.
+\end{equation}
+\end{itemize}
+
+\section{Intro2}
+In this note we establish the following results:
+\begin{enumerate}
+\item
+The regularor of $\tilde A$
+grows as $O (\ln^{n-1} (p)) $;
+
+\item
+If $n$ is prime,
+the index of the centralizers
+$ [Z_{\SLn}(\tilde A) : Z_{\SLp}(\tilde A)] < C$
+is bounded by some constant $C$ and does not depend on $p$.
+\end{enumerate}
+This might be interesting.
+
+\section{Regulators}
+We mimic the proofs from a Keigh Conrad's write-up on the Dirichlet unit theorem \cite{conraddirichlet}.
+
+There are more general estimates due to Duke \cite{duke2003extreme} (see Section 3 Proposition 1).
+
+Let $K \eqdef \Q(\alpha)$ be the number field obtain by attaching the root of $f(t)$ to $\Q$.
+\begin{Prop}
+$K / \Q$ is a totally real number field of degree $n$.
+\end{Prop}
+
+\begin{proof}
+First, $f(t)$ is irreducible as it has no rational roots: 
+\begin{gather}
+f(1) = 1 - (1 + p) (1 + 2p) \ldots (1 + (n - 1) p) \ne 0,\\
+f(-1) = (-1)^n - (1 - p) (1 - 2p) \ldots (1 - (n - 1) p ) \ne 0.
+\end{gather}
+
+Let's compute the roots of $f(t)$ approximately and show they are real.
+
+Notice that for $|t| \ll 1$ all but one roots of $f(t)$ are close to the roots of 
+\[
+(1 + pt) (1 + 2pt) \ldots (1 + (n - 1) p t)
+,\] 
+them being $p^{-1} , \frac{1}{2} p^{-1} , \ldots, \frac{1}{(n-1) } p^{-1} $.
+
+As all but one roots are real, the remaining one is also real.
+\end{proof}
+
+The next goal is to tackle the unit group $\O_K^*$ of $K$.
+
+The Dirichlet unit theorem implies that $\O_K^*$ is an abelian group of rank $n-1$ (there might be some finite torsion of roots of unity.  It is bounded universally in $p$).
+% depending only on $n$
+% as the degree of the roots of unity is the totient function which grows as $n \to \infty$.
+A basis of $\O_K^*$ modulo roots of unity is called a system of fundamental units.
+Finding a system of fundamental units in general is a complicated task.
+
+Instead, we shall find a finite index subgroup $U \subset \O_K^*$. 
+% \red{maybe order $\O \subset \O_K$ instead?}
+Then we shall bound above the regulator of $K$ with the regulator of $U$.
+
+Moreover, using a Silverman's lower bound on the regulator of $K$ we bound above the index $[\O_K^*: U] < C_1$ with $C_1$ independent on $p$.
+% \red{maybe say uniformly in $p$ instead?}
+
+Define a subgroup 
+\[
+	U = (1 + p \alpha)^{\mathbb{Z}} \ldots (1 + (n-1) p \alpha)^{\mathbb{Z}} \subset \O_K^*
+.\] 
+\begin{Prop}
+$U \subset \O_K^*$ is a finite index subgroup.
+The regulator $\Reg_U$ is bounded above by $C_2 \log(p)^{n-1}$ where $C_2$ is a constant independent on $p$.
+\end{Prop}
+
+\begin{proof}
+We have
+\begin{gather}
+\alpha^n - (1 + p\alpha) (1 + 2p\alpha) \ldots (1 + (n - 1) p\alpha) + 1 = 1  ,
+\\
+(1 + p\alpha) (1 + 2p\alpha) \ldots (1 + (n - 1) p\alpha) = \alpha^n.
+\end{gather}
+It shows that $1 + p \alpha, \ldots, 1 + (n - 1) p \alpha$ are indeed units.
+
+Let $\alpha_1 < \ldots < \alpha_{n-1} < 0 $ be the negative roots of $f$.
+We shall compute them approximately.
+\begin{Lem}
+\[
+\alpha_k = \frac{-1}{k} p^{-1} + \frac{(-1)^{k + 1} }{ k\cdot  k! (n - k + 1)!} p^{-n-1}  + O(p^{-2 n - 1})
+.\] 
+\end{Lem}
+\begin{proof}
+In the first approximation $\alpha_k = \frac{-1}{k} p^{-1} + o(p^{-1})$.
+Let's compute the residue term more persisely. 
+Let $\alpha_k = \frac{-1}{k} p^{-1} ( 1 + \epsilon )$.
+Substituting to $f(t)$, we get
+\begin{gather*}
+(1 + p\alpha_k) (1 + 2p\alpha_k) \ldots (1 + (n - 1) p\alpha_k) = \alpha_k^n, \\
+- (- k p\alpha_k - k) (- 2 k p\alpha_k - k)  \ldots (- (n - 1) k p\alpha_k - k)  = (- k \alpha_k) ^n, \\
+- (1 - \epsilon - k) (2 - 2 \epsilon - k)  \ldots ( (n-1) - (n-1) \epsilon - k)  = p^{-n} (1 + \epsilon)^n, \\
+- (1 - k) (2 - k) \ldots (- 1) k \epsilon 1 \ldots ( (n-1) - k) + O ( \epsilon^2)
+= p^{-n} + O( \epsilon p^{-n} ) , \\
+(-1)^k k! ( n - k - 1)! \epsilon + O ( \epsilon^2)
+= p^{-n} + O( \epsilon p^{-n} ) , \\
+\epsilon = \frac{ (-1)^k p^{-n} }{ k! ( n - k - 1)! }  + O( \epsilon p^{-n} ) + O ( \epsilon^2), \\
+\epsilon = \frac{ (-1)^k p^{-n} }{ k! ( n - k - 1)! }  + O( p^{-2n} ).
+\end{gather*}
+\end{proof}
+
+By the definition of the regulator we have
+\begin{multline}
+\Reg_U = 
+\left|
+\begin{array}{cccc}
+\ln |1 + p \alpha_1| 	& \ln |1 + p \alpha_2| & \ldots & \ln |1 + p \alpha_{n-1}| \\
+\ln |1 + 2 p \alpha_1| 	& \ln |1 + 2 p \alpha_2| & \ldots &  \ln |1 + 2 p \alpha_{n-1}| \\
+\vdots 	& \vdots & \ddots &  \vdots \\
+\ln |1 + (n-1) p \alpha_1| 	& \ln |1 + (n-1) p \alpha_2| & \ldots &  \ln |1 + (n-1) p \alpha_{n-1}| 
+\end{array} 
+\right|
+\\
+\approx
+\left|
+\begin{array}{cccc}
+\ln \left( \frac{ p^{-n}}{n!}  \right)  	& \ln \left( 1 - \frac{1}{2} \right)  & \ldots &  \ln \left( 1 - \frac{1}{n - 1} \right)  \\
+\ln |1 - 2| 	& \ln \left( \frac{p^{-n} }{2\cdot  2! (n-1)!} \right)  & \ldots &  \ln \left( 1 - \frac{2}{n-1} \right)  \\
+\vdots 	& \vdots & \ddots &  \vdots \\
+\ln |1 - (n-1) | 	& \ln |1 - \frac{n-1}2| & \ldots &  \ln \left( \frac{p^{-n}}{(n-1)\cdot (n-1)! 2!} \right) 
+\end{array} 
+\right| \\
+\approx n^{n-1} \log(p)^{n-1} 
+.\end{multline}
+\end{proof}
+From the proof of the Proposition above one finds that the regulator of $U$ $ \Reg_U > 0 $ is positive for big enough prime $p$.
+Hence the generators $(1 + p \alpha), \ldots, (1 + (n-1) p)$ of $U$ are multiplicatively independent, and $U \subset \O_K^*$ is a finite index subgroup.
+
+We shall bound the index $[\O_K^* : U]$ from above via the following
+\begin{Thm}[Silverman, \cite{silverman1984inequality}]
+Let $K$ be a number field of degree $d$ with regulator $R$, and absolute discriminant $D_K$. 
+Let $r(K)$ be the rank of the unit group in $K$, and let $p(K)$ be the maximum of $r(k)$ as $k$ ranges over proper subfields of $K$. 
+We prove 
+\[
+R_K > c_d (\log \gamma_d  D_K )^{r(K) - p(K)} 
+\] 
+for constants $c_d, \gamma_d$ depending only on $d$. 
+\end{Thm}
+\red{It seems it's actually due to 
+Robert Remak in
+"Über Größenbeziehungen zwischen Diskriminante
+und Regulator eines algebraischen Zahlkörpers" }
+
+\begin{Prop}
+Let $n$ be a prime.
+Then the index $[\O_K^* : U] < C_3$ is bounded from above by a constant $C_3$ independent on $p$.
+\end{Prop}
+\begin{proof}
+The discriminant of $K$ is the discriminant of the polynomial $f(t)$, which is approximately
+\[
+	\disc_K = \disc_f = O( p^{2 (n - 1)^2} p^{-(n - 1) (n - 2)} ) = O ( p^{n (n - 1) } )
+.\] 
+More precisely,
+\[
+	\disc_K = \disc_f \approx 
+	1!^2 \ldots (n-1)!^2 p^{n (n - 1) }
+.\]
+
+Given that $n$ is prime and so there are no nontrivial subfield of $K$ the Silverman's theorem implies 
+$$\Reg_K > C_4 (\log \disc_K)^{n-1} \approx n^{n-1} (n-1)^{n-1} C_4 \log(p)^{n-1}  > C_5\log(p)^{n-1}.$$
+
+It is easy to see that 
+\[
+\Reg_U = [\O_K^* : U] \Reg_K
+.\] 
+Thus we have
+\[
+ C_2 \log(p)^{n-1} > \Reg_U = 
+[\O_K^* : U] \Reg_K >
+[\O_K^* : U] C_5 \log(p)^{n-1}
+.\] 
+It follows that the index is bounded from above
+\[
+[\O_K^* : U] < \frac{C_2}{C_5}
+,\] 
+as desired.
+\end{proof}
+
+\section{Centralizers}
+\begin{Prop}
+Let $n$ be a prime.  
+Then the index of the centralizers
+\[
+	[Z_{\SLn}(\tilde A) : Z_{\SLp}(\tilde A)] < C_3
+\] 
+is bounded from above by a constant independent of $p$.
+\end{Prop}
+\begin{proof}
+	We shall exploit the properties of the previously introduced number field $K = \mathbb{Q}(\alpha)$, where $\alpha$ is a root of $f(t) = t^n - (1 + pt) \ldots (1 + (n-1) pt)$.
+
+Since $A$ is regular, 
+% over $\mathbb{Z}$
+there is an isomorphism of $\mathbb{Z}$-algebras
+
+\[
+\Phi: \mathbb{Z}[A] 
+\mathop{\longrightarrow}^{\sim}
+\mathbb{Z}[\alpha], A \mapsto \alpha
+.\]
+
+It identifies the stabilizer of $\tilde A$ in $\SLn$ with the multiplicative group of the order $\mathbb{Z}[\alpha] \subset \O_K$.
+\[
+Z_{\SLn}(\tilde A) 
+= Z_{\SLn}(A) 
+= \mathbb{Z}[A]^*
+\mathop{\simeq }^{\Phi} 
+\mathbb{Z}[\alpha]^*
+.\]
+
+The isomorphism $\Phi$ identifies
+the subgroup $U \subset \mathbb{Z}[\alpha]^*$ with
+$$
+\Phi^{-1} (U) = \tilde A^{\mathbb{Z}} (2 \tilde A - \Id )^{\mathbb{Z}} \ldots   ((n-1) \tilde A - (n-2) \Id )^{\mathbb{Z}} \subset Z_{\SLp}(\tilde A) 
+.$$
+
+Thus, there is a chain of inclusions
+\[
+\Phi^{-1} (U) \subset Z_{\SLp}(\tilde A) 
+\subset Z_{\SLn}(\tilde A) = \Phi^{-1} \mathbb{Z}[\alpha]^*
+.\] 
+Therefore, there is a bound
+\[
+[Z_{\SLn}(\tilde A) : Z_{\SLp}(\tilde A) ]
+\le
+[\mathbb{Z}[\alpha]^*, U]
+\le 
+[\O_K^*, U] < C_3
+.\] 
+\end{proof}
+
 \bibliography{mybib}

### 2024-10-02 draft_systoles.tex
added the stiefel_whitney class

@@
 \item
-The regularor of $\tilde A$
+The regulator of $\tilde A$
 grows as $\approx 3 \ln^2(p)$.
@@
 
+\newpage
+\section{Computing (co)homology}
+Christophe Soulé \cite{soule1978}
+computed the cohomology of $\SL$
+to be
+\begin{table}[h!]
+	\centering
+\begin{tabular}{|c|c|c|c|c|c|c}
+	\hline
+	$n$        & $0$                               & $1$                               & $2$                               & $3$                               & $4$  & \ldots\\
+\hline 
+	$H^n(\SL, \mathbb{Z})$ & $\mathbb{Z}$ & $0$ & $0$ & $(\mathbb{Z}/2\mathbb{Z})^{2} $ & $(\mathbb{Z}/3\mathbb{Z})^{2} \oplus (\mathbb{Z}/4\mathbb{Z})^{2} $ & \ldots \\
+\hline
+\end{tabular}.
+\end{table}
+
+The universal coefficient formula gives us the homology.
+\begin{table}[h!]
+	\centering
+\begin{tabular}{|c|c|c|c|c|c}
+	\hline
+	$n$        & $0$                               & $1$                               & $2$                               & $3$                               & \ldots\\
+\hline 
+	$H_n(\SL, \mathbb{Z})$ & $\mathbb{Z}$ & $0$ &  $(\mathbb{Z}/2\mathbb{Z})^{2} $ & $(\mathbb{Z}/3\mathbb{Z})^{2} \oplus (\mathbb{Z}/4\mathbb{Z})^{2} $ & \ldots \\
+\hline
+\end{tabular}
+\end{table}
+
+Recall the matrices 
+\[
+A = 
+\begin{pmatrix}
+0 & 0 & 1 \\
+1 & 0 & 0 \\
+0 & 1 & -p^2
+\end{pmatrix}
+,
+B = \Id + p A =
+\begin{pmatrix}
+1 & 0 & p \\
+p & 1 & 0 \\
+0 & p & 1 - p^3
+\end{pmatrix}
+.\] 
+We denote $\tilde A$ by $B$ for convenience.
+
+As $A, B $ commute they define a map
+\[
+f: \mathbb{Z}^2 \to \SL, 
+\] 
+\[
+(k, l) \mapsto A^k B^l
+.\] 
+\fbox{
+\parbox{\textwidth}{
+The goal for us it to understand the map 
+\[
+H_2(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z} \to 
+H_2(\SL, \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} 
+.\] 
+}}
+
+Consider the natural embedding 
+\[
+i: \SL \embeds {\mathbf{SL}_3(\mathbb{R})}
+.\] 
+\begin{Prop}
+The pullback 
+\[
+i^*: H^2(B\mathbf{SL}_3(\mathbb{R}), \mathbb{Z}/ 2 \mathbb{Z}) \cong \mathbb{Z} / 2 \mathbb{Z} \to 
+H^2(\SL, \mathbb{Z}/ 2 \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2}
+\] 
+is an embedding.
+\end{Prop}
+\begin{proof}
+Consider a pair of matrices
+\[
+C_1 = 
+\begin{pmatrix}
+-1 & 0 & 0 \\
+0 & -1 & 0 \\
+0 & 0 & 1
+\end{pmatrix}
+,
+C_2 =
+\begin{pmatrix}
+1 & 0 & 0 \\
+0 & -1 & 0 \\
+0 & 0 & -1
+\end{pmatrix} \in \SL
+.\] 
+They give rise to a principle $\SL$ bundle $E$ on $\mathbb{T}^2$ with non-zero $w_2$, the second Stiefel-Whitney class.
+
+Indeed, by the splitting principle we have
+\[
+w(E) = w(E_1) w(E_2) w (E_3)
+= (1 + x) (1 + x + y) (1 + y) 
+= 1 
++ 2 (x + y) 
++ (x^2 + xy + xy + xy + y^2) 
++ x ( x + y) y 
+= 1 + xy
+,\] 
+where $x, y$ denote the generators of $H^1(\mathbb{T}^2, \mathbb{Z}/ 2 \mathbb{Z})$.
+Therefore the composition 
+\[
+H^2(B\mathbf{SL}_3(\mathbb{R}), \mathbb{Z}/ 2 \mathbb{Z}) \cong \mathbb{Z} / 2 \mathbb{Z} \to 
+H^2(\SL, \mathbb{Z}/ 2 \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2}
+\to
+H^2(\mathbb{T}^2, \mathbb{Z}/ 2 \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})
+\] 
+is nonzero and thus the first map is an embedding.
+\end{proof}
+Thus $w_2$ is one of the generators of the second cohomology of $\SL$.
+Consider the principal $\SL$-bundle $E$ on $\mathbb{T}^2$ associated with $A, B$.
+\begin{Prop}
+	$w_2(E) = 0$.
+\end{Prop}
+\begin{proof}
+By the splitting principle
+\[
+w(E) = w(E_1) w(E_2) w (E_3)
+= (1 + x + y) (1 + x + y) (1) 
+= 1
+.\]
+\end{proof}
+There are perfect pairings
+\[
+H_2(\SL, \mathbb{Z}) \times 
+H^2(\SL, \mathbb{Z}/2 \mathbb{Z}) \to \mathbb{Z}/ 2 \mathbb{Z}
+,\] 
+\[
+H_2(\mathbb{T}^2, \mathbb{Z}) \times 
+H^2(\mathbb{T}^2, \mathbb{Z}/2 \mathbb{Z}) \to \mathbb{Z}/ 2 \mathbb{Z}
+.\] 
+Combining them, we get that the map
+\[
+H_2(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z} \to 
+H_2(\SL, \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} 
+\] 
+has one coordinate equal to zero.
+
+\newpage
 \section{From $\SL$ to $\SLn$}

### 2024-10-02 draft_systoles.tex
computed push in H_2 for SL_3(Z)

--- 2024-07-10 draft_systoles.tex
@@
 \DeclareMathOperator{\disc}{disc}        % discriminant
+\DeclareMathOperator{\Hom}{Hom}        % Hom
+\DeclareMathOperator{\Aut}{Aut}        % Aut
 
@@
 \begin{Rem}
+	\label{rem_roots_estimate}
 A computation shows that for $p = 100$ we have
@@
 
+\section{Computing (co)homology}
+Christophe Soulé \cite{soule1978}
+computed the cohomology of $\SL$
+to be
+\begin{table}[h!]
+	\centering
+\begin{tabular}{|c|c|c|c|c|c|c}
+	\hline
+	$n$        & $0$                               & $1$                               & $2$                               & $3$                               & $4$  & \ldots\\
+\hline 
+	$H^n(\SL, \mathbb{Z})$ & $\mathbb{Z}$ & $0$ & $0$ & $(\mathbb{Z}/2\mathbb{Z})^{2} $ & $(\mathbb{Z}/3\mathbb{Z})^{2} \oplus (\mathbb{Z}/4\mathbb{Z})^{2} $ & \ldots \\
+\hline
+\end{tabular}.
+\end{table}
+
+The universal coefficient formula gives us the homology.
+\begin{table}[h!]
+	\centering
+\begin{tabular}{|c|c|c|c|c|c}
+	\hline
+	$n$        & $0$                               & $1$                               & $2$                               & $3$                               & \ldots\\
+\hline 
+	$H_n(\SL, \mathbb{Z})$ & $\mathbb{Z}$ & $0$ &  $(\mathbb{Z}/2\mathbb{Z})^{2} $ & $(\mathbb{Z}/3\mathbb{Z})^{2} \oplus (\mathbb{Z}/4\mathbb{Z})^{2} $ & \ldots \\
+\hline
+\end{tabular}
+\end{table}
+
+Recall the matrices 
+\[
+A = 
+\begin{pmatrix}
+0 & 0 & 1 \\
+1 & 0 & 0 \\
+0 & 1 & -p^2
+\end{pmatrix}
+,
+B = \Id + p A =
+\begin{pmatrix}
+1 & 0 & p \\
+p & 1 & 0 \\
+0 & p & 1 - p^3
+\end{pmatrix}
+.\] 
+We denote $\tilde A$ by $B$ for convenience.
+
+As $A, B $ commute they define a map
+\[
+f: \mathbb{Z}^2 \to \SL, 
+\] 
+\[
+(k, l) \mapsto A^k B^l
+.\]
+
+\begin{Thm}
+	\label{thm_push_homology_vanishes}
+The pushforward map in the second homology 
+\[
+f_*:
+H_2(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z} \to 
+H_2(\SL, \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} 
+\] 
+is zero.
+\end{Thm}
+The goal of this section is to prove this theorem.
+
+The first observation is that one can pass from the homology to the cohomology with coefficients $\mathbb{Z}/ 2 \mathbb{Z}$.
+
+\begin{Lem}
+	\label{lemma_push_vanishes_iff_pull_vanishes}
+The pushforward
+\[
+f_*:
+H_2(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z} \to 
+H_2(\SL, \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} 
+\] 
+is zero 
+iff
+the pullback
+\[
+f^*:
+H^2(\SL,  \mathbb{Z}/2 \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} \to
+H^2(\mathbb{Z}^2, \mathbb{Z}/2 \mathbb{Z}) \cong \mathbb{Z}/2 \mathbb{Z}
+\] 
+is zero.
+\end{Lem}
+\begin{proof}
+As the first homology 
+\[
+H_1(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z}^2, 
+H_1(\SL, \mathbb{Z}) \cong 0
+\] 
+are torsion-free, the universal coefficien formula gives perfect $\mathbb{Z}/2\mathbb{Z}$-pairings between homology and cohomology
+\[
+H^2(\mathbb{Z}^2, \mathbb{Z}/2 \mathbb{Z}) = \Hom(H_2(\mathbb{Z}^2, \mathbb{Z}), \mathbb{Z}/2 \mathbb{Z})
+,\] 
+\[
+H^2(\SL, \mathbb{Z}/2 \mathbb{Z}) = \Hom(H_2(\SL, \mathbb{Z}), \mathbb{Z}/2 \mathbb{Z})
+.\] 
+Therefore the pullback in cohomology is the transposed map of the pushforward in homology and one is zero iff the other is zero.
+\end{proof}
+
+Next, let us find a nicer description for $H^2(\SL, \mathbb{Z}/2 \mathbb{Z})$.
+
+\begin{Cla}
+\begin{itemize}
+	\item The second cohomology $H^2(G, M)$ classify the extensions $0 \to M \to \tilde G \to G \to 0$;
+	\item In particular, $H^2(G, \mathbb{Z}/2\mathbb{Z})$ correspond to the extensions of degree $2$;
+\item Given an extension $0 \to M \to \tilde G \to G \to 0$ and a principle $G$-bundle over $X$ we get its class in $H^2(X, M)$ via the boundary (in other words, Bockstein) homomorphism $H^1(X, G) \to H^2(X, M)$.
+In details, the Bockstein goes by taking the $1$-cocycle defining the $G-$bundle to $\tilde G$ and apply the Cech differential;
+	\item $H^2( \mathbf{SL}_3(\mathbb{R}), \mathbb{Z}/ 2 \mathbb{Z}) \cong
+	H^2( \mathbf{SL}_3(\mathbb{F}_2), \mathbb{Z}/ 2 \mathbb{Z}) \cong \mathbb{Z}/2\mathbb{Z}$;
+\item The corresponding unique extensions of degree two are $\widetilde{\mathbf{SL}_3}(\mathbb{R})$ and $\mathbf{SL}_2(\mathbb{F}_7)$.
+	We make use of the exceptional isomorphism $\mathbf{PSL}_2(\mathbb{F}_7) \cong \mathbf{SL}_3(\mathbb{F}_2)$, see, for example, \cite{merzon_psl_2_7_sl_3_2}.
+\item If $E$ is a principle $\mathbf{SL}_3(\mathbb{R})$-bundle on $X$ the corresponding class in $H^2(X, \mathbb{Z}/ 2 \mathbb{Z})$ is called $w_2(E)$, the second Stiefel-Whitney class;
+\item If $E$ is a principle $\mathbf{SL}_3(\mathbb{F}_2)$-bundle on $X$ let us denote the corresponding class in $H^2(X, \mathbb{Z}/ 2 \mathbb{Z})$ by $u_2(E)$.
+\end{itemize}
+\end{Cla}
+This is all very well known.
+For the cohomology of $\mathbf{SL}_3(\mathbb{F}_2)$ see, for example, \cite{adem2006lecturescohomologyfinitegroups}.
+
+Now, consider the natural embeddings
+\[
+i_1: \SL \embeds {\mathbf{SL}_3(\mathbb{R})}
+, \] 
+\[
+i_2: \SL \embeds {\mathbf{SL}_3(\mathbb{F}_2)}
+.\]
+
+\begin{Prop}
+The pullback 
+\[
+i_1^* \oplus i_2^*: 
+H^2(B\mathbf{SL}_3(\mathbb{R}), \mathbb{Z}/ 2 \mathbb{Z}) \oplus
+H^2(B\mathbf{SL}_3(\mathbb{F}_2), \mathbb{Z}/ 2 \mathbb{Z}) \cong (\mathbb{Z} / 2 \mathbb{Z})^2 \to 
+H^2(\SL, \mathbb{Z}/ 2 \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2}
+\] 
+is an isomorphism.
+\end{Prop}
+
+\begin{proof}
+It is enough to show that a principal $\SL$ bundle $E$ on $\mathbb{T}^2$ can have an arbitrary combination of $(w_2(E), u_2(E)) \in (\mathbb{Z}/ 2\mathbb{Z})^2$.
+
+Plan: we shall construct principal bundles $E_1, E_2$ on $\mathbb{T}^2$ with $(w_2(E_1), u_2(E_1)) = (1, 0)$ and $u_2(E_2) = 1$.
+This will prove the proposition.
+
+The first bundle $E_1$ is defined via the monodromy matrices $C_1, C_2$
+\[
+C_1 = 
+\begin{pmatrix}
+-1 & 0 & 0 \\
+0 & -1 & 0 \\
+0 & 0 & 1
+\end{pmatrix}
+,
+C_2 =
+\begin{pmatrix}
+1 & 0 & 0 \\
+0 & -1 & 0 \\
+0 & 0 & -1
+\end{pmatrix} \in \SL
+.\] 
+By the splitting principle we have
+\[
+w(E_1)
+= (1 + x) (1 + x + y) (1 + y) 
+= 1 + xy
+,\] 
+where $x, y$ denote the generators of $H^1(\mathbb{T}^2, \mathbb{Z}/ 2 \mathbb{Z})$.
+Therefore $w_2(E_1) = 1$.
+As modulo $2$ the matrices $C_1, C_2$ are the identity the bundle $E_1$ mod $2$ is trivial; thus $u_2(E_1) = 0$.
+
+Define $D_1' = 
+\left( \begin{smallmatrix}0&6\\1&0\end{smallmatrix}\right) 
+, D_2' =
+\left( \begin{smallmatrix}2&3\\3&5\end{smallmatrix}\right)
+\in \mathbf{SL}_2(\mathbb{F}_7)$.
+Project the matrices to 
+$\in \mathbf{PSL}_2(\mathbb{F}_7) \cong \in \mathbf{SL}_3(\mathbb{F}_2) $.
+Pick a pair of lifts $D_1, D_2 \in \mathbf{SL}_3(\mathbb{Z})$.
+
+Define the bundle $E_2$ corresponding to the monodromy matrices $D_1, D_2$. 
+We claim that $u_2(E_2) = 1$.
+
+Indeed, we shall compute the Bockstein morphism $H^1(\mathbb{T}^2, \mathbf{SL}_3(\mathbb{F}_2) \to H^2(\mathbb{T}^2, \mathbb{Z}/ 2 \mathbb{Z})$ by lifting the cocycles.
+
+One checks that $D_1' D_2' = - D_2' D_1'$; thus the image of the Bockstein map is nontrivial.
+
+Thus we have showed that $E_1, E_2$ have the right characteristic classes and we are done.
+\end{proof}
+\begin{proof}[Proof of the Theorem \ref{thm_push_homology_vanishes}]
+By Lemma \ref{lemma_push_vanishes_iff_pull_vanishes} we shall show that the pullback map in second $\mathbb{Z}/ 2 \mathbb{Z}$-cohomology vanishes.
+
+Consider the principal $\SL$-bundle $E$ on $\mathbb{T}^2$ associated with $A, B$.
+We should show that $w_2(E) = 0$ and $u_2(E) = 0$.
+
+By the splitting principle
+using that the roots of the characteristic polynomial 
+\ref{rem_roots_estimate} 
+satisfy \[
+\alpha_1, 1 + p \alpha_1 < 0, 
+\alpha_2, 1 + p \alpha_2 < 0, 
+\alpha_3, 1 + p \alpha_3 > 0 
+\] 
+we get the Stiefel-Whitney class to be
+\[
+w(E) = w(E_1) w(E_2) w (E_3)
+= (1 + x + y) (1 + x + y) (1) 
+= 1
+,\]
+and thus $w_2(E) = 0$ indeed does vanish.
+
+Modulo $2$ one easily spots that $B = A^5 \pmod{2}$, thus their lifts to $\mathbf{SL}_2(\mathbf{F}_7)$ do commute and the image of the Bockstain morphism vanishes; thus $u_2(E) = 0$.
+% the explanation is that we should consider SL_3(F_2) in the automorphisms of F_8 = F_2[t]/t^3 + t^2 + 1 as a linear space over F_2; then A is multiplication by t and B is a multiplication by 1 + t = t^5.
+\end{proof}
+
+\newpage
 \section{From $\SL$ to $\SLn$}
@@
 \bibliography{mybib}
-\bibliographystyle{plain}
+\bibliographystyle{alpha}
 

### 2024-10-03 draft_systoles.tex
Add explicit integer lifts
minor corrections

@@
 
-Now, consider the natural embeddings
+Now, consider the natural mappings
 \[
-i_1: \SL \embeds {\mathbf{SL}_3(\mathbb{R})}
+i: \SL \embeds {\mathbf{SL}_3(\mathbb{R})}
 , \] 
 \[
-i_2: \SL \embeds {\mathbf{SL}_3(\mathbb{F}_2)}
+\pi: \SL \embeds {\mathbf{SL}_3(\mathbb{F}_2)}
 .\]
@@
 \[
-i_1^* \oplus i_2^*: 
+i^* \oplus \pi^*: 
 H^2(B\mathbf{SL}_3(\mathbb{R}), \mathbb{Z}/ 2 \mathbb{Z}) \oplus
@@
 $\in \mathbf{PSL}_2(\mathbb{F}_7) \cong \in \mathbf{SL}_3(\mathbb{F}_2) $.
-Pick a pair of lifts $D_1, D_2 \in \mathbf{SL}_3(\mathbb{Z})$.
+Pick a pair of commuting lifts $D_1, D_2 \in \mathbf{SL}_3(\mathbb{Z})$.
+One can take them to be 
+\begin{equation}
+D_1 = 
+\begin{pmatrix}
+1 & 1 & 0 \\
+0 & 1 & 0 \\
+0 & 1 & 1
+\end{pmatrix}, 
+D_2 =
+\begin{pmatrix}
+1 & 0 & 0 \\
+0 & 1 & 0 \\
+0 & 1 & 1
+\end{pmatrix}
+.\end{equation}
 
@@
 \begin{proof}[Proof of the Theorem \ref{thm_push_homology_vanishes}]
-By Lemma \ref{lemma_push_vanishes_iff_pull_vanishes} we shall show that the pullback map in second $\mathbb{Z}/ 2 \mathbb{Z}$-cohomology vanishes.
+In view of the Lemma \ref{lemma_push_vanishes_iff_pull_vanishes} we shall show that the pullback map in second $\mathbb{Z}/ 2 \mathbb{Z}$-cohomology vanishes.
 

### 2024-10-03 draft_systoles.tex
minor misprints corrections

@@
 \section{Regulators}
-We mimic the proofs from a Keigh Conrad's write-up on the Dirichlet unit theorem \cite{conraddirichlet}.
+We mimic the proofs from a Keith Conrad's write-up on the Dirichlet unit theorem \cite{conraddirichlet}.
 
@@
 	\item The second cohomology $H^2(G, M)$ classify the extensions $0 \to M \to \tilde G \to G \to 0$;
-	\item In particular, $H^2(G, \mathbb{Z}/2\mathbb{Z})$ correspond to the extensions of degree $2$;
+	\item In particular, $H^2(G, \mathbb{Z}/2\mathbb{Z})$ correspond to the extensions of index two;
 \item Given an extension $0 \to M \to \tilde G \to G \to 0$ and a principle $G$-bundle over $X$ we get its class in $H^2(X, M)$ via the boundary (in other words, Bockstein) homomorphism $H^1(X, G) \to H^2(X, M)$.
@@
 	H^2( \mathbf{SL}_3(\mathbb{F}_2), \mathbb{Z}/ 2 \mathbb{Z}) \cong \mathbb{Z}/2\mathbb{Z}$;
-\item The corresponding unique extensions of degree two are $\widetilde{\mathbf{SL}_3}(\mathbb{R})$ and $\mathbf{SL}_2(\mathbb{F}_7)$.
+\item The corresponding unique extensions of index two are $\widetilde{\mathbf{SL}_3}(\mathbb{R})$ and $\mathbf{SL}_2(\mathbb{F}_7)$.
 	We make use of the exceptional isomorphism $\mathbf{PSL}_2(\mathbb{F}_7) \cong \mathbf{SL}_3(\mathbb{F}_2)$, see, for example, \cite{merzon_psl_2_7_sl_3_2}.
@@
 \[
-\pi: \SL \embeds {\mathbf{SL}_3(\mathbb{F}_2)}
+\pi: \SL \projects {\mathbf{SL}_3(\mathbb{F}_2)}
 .\]
@@
 Project the matrices to 
-$\in \mathbf{PSL}_2(\mathbb{F}_7) \cong \in \mathbf{SL}_3(\mathbb{F}_2) $.
+$\in \mathbf{PSL}_2(\mathbb{F}_7) \simeq \mathbf{SL}_3(\mathbb{F}_2) $.
 Pick a pair of commuting lifts $D_1, D_2 \in \mathbf{SL}_3(\mathbb{Z})$.
@@
 
-By the splitting principle
+Diagonalizing $A, B$ over reals by the splitting principle
 using that the roots of the characteristic polynomial 
 \ref{rem_roots_estimate} 
-satisfy \[
+satisfy 
+\[
 \alpha_1, 1 + p \alpha_1 < 0, 
+\] 
+\[
 \alpha_2, 1 + p \alpha_2 < 0, 
-\alpha_3, 1 + p \alpha_3 > 0 
+\] 
+\[
+\alpha_3, 1 + p \alpha_3 > 0 .
 \] 
@@
 \end{proof}
+\begin{Rem}
+Similar computations show that the pushforward in second homology of $\mathbb{Z}^2 \to \SL$ defined via $A^3, B$ vanishes, too.
+\end{Rem}
 
